---
title: 'Demo of `abm()` function in abmneo'
output: pdf_document
author: Sasha D. Hafner
date: "`r format(Sys.time(), '%d %B, %Y %H:%M')`"
---

# Overview
This demo shows:

# Prep

```{r}
devtools::load_all()
```

# Default input parameters

We need to set some parameters, first management parameters.
Here are some default sets, not always with plausible values.

Now pared-down mng_pars, only those that apply to both types ().

```{r}
mng_pars <- list(
  storage_depth = 2,     
  area = 100,              
  temp_C = 20,
  resid_enrich = 0.2 
)
```

Next substrate parameters, a new argument.
This defines substrates.
We could have any number with any names.
Note that hydrolysis uses CTM again (like anything here, that could be changed).

```{r}
sub_pars <- list(
  subs = c('VSd'),
  T_opt_hyd = c(VSd = 60),
  T_min_hyd = c(VSd = 0),
  T_max_hyd = c(VSd = 90),
  hydrol_opt = c(VSd = 0.1),
  sub_fresh = c(VSd = 50),
  sub_init = c(VSd = 50)
)
```

Microbial parameters are similar to other ABM versions, but inhibition is set separately now (and not included in this simple example).
But now we have microbial stoichiometry instead of the "special cases" that greatly complicate the code.
The ks par is now in a matrix, which could have more than one row (e.g., with sulfate reduction).

```{r}
mstoich <- matrix(
  c(-1, -1, -1, 1, 1, 1),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'CH4'),
    c('m0', 'm1', 'm2')
  )
)

mstoich

ksmat <- matrix(
  c(0.5, 0.5, 0.5),
  nrow = 1,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH'),
    c('m0', 'm1', 'm2')
  )
)

ksmat

```



```{r}
grp_pars <- list(
  grps = c('m0', 'm1', 'm2'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41),
  T_max = c(m0 = 25, m1 = 25, m2 = 38),
  ksmat = ksmat,
  mstoich = mstoich
)
```

The `dd_rate` parameter is for "death and decay".

These last two arguments are similar to other versions.
VFA is hard-coded as an intermediate and has its own elements.
A name is needed, and is `CH3COOH` here (and hard-coded for now).

```{r}
man_pars <- list(VFA_fresh = c(CH3COOH = 2), VFA_init = c(CH3COOH = 2), pH = 7, dens = 1000)
```

The chem_pars list now specifies which components are emitted gases and which are otherwise solutes.
(All substrates are enriched--perhaps this should be an input too.)

```{r}
chem_pars <- list(COD_conv = c(CH4 = 3.01), gases = c('CH4'))
```

# 1. Slurry mass series versus regular inputs
When it comes to how slurry level and the fill/empty schedule is entered, we now distinguish between *regular* inputs (as in repeating/consistent, as in earlier versions of `abm()`) and *series* inputs, where the schedule is inferred from time series data on slurry mass (what was called time-variable in earlier versions of `abm()`).
These are specified in the `structure` argument (to be renamed. . .  maybe simply `mass`???).
Mass dat for structure only has time and slurry_mass.
It is no longer a part of `var_pars`--the two are now independent (externally).

```{r}
mass_series <- data.frame(
  time = c(0, 1, 10, 15, 30), 
  slurry_mass = c(1000, 7000, 2000, 1500, 3000)
)

structure1 <- list(
  type = 'series',
  dat = mass_series
)

plot(slurry_mass ~ time, data = structure1$dat, type = 'o')
```

```{r}
devtools::load_all()
out1 <- abm(
  structure = structure1,
  50,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars
)
names(out1)
```
traceback()
Q

No `var_pars` needed.
And the fixed inputs like `slurry_prod_rate` that were ignored in these cases are no longer expected.

```{r}
plot(slurry_mass ~ time, data = out1, type = 'l')
plot(CH4_emis_rate ~ time, data = out1, type = 'l')
plot(CH4 ~ time, data = out1, type = 'l')
plot(m0 ~ time, data = out1, type = 'l')
plot(CH3COOH ~ time, data = out1, type = 'l')
plot(VSd_conc ~ time, data = out1, type = 'l')
plot(m0_conc ~ time, data = out1, type = 'l')
plot(COD_load ~ time, data = out1, type = 'l')
```

With regular inputs, `slurry_prod_rate` etc. are now explicitly given.

```{r}
structure2 <- list(
  type = 'regular',
  slurry_prod_rate = 1E4, 
  slurry_mass = 1E4,     
  resid_depth = 0.1,      
  empty_int = 10,          
  wash_water = 100,            
  wash_int = NA,
  rest_d = 1
)
```

```{r}
devtools::load_all()
out2 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars
)
```
Q

```{r}
out2$slurry_mass
out2$slurry_mass_eff
```

```{r}
plot(slurry_mass ~ time, data = out2, type = 'l')
plot(CH4_emis_rate ~ time, data = out2, type = 'l')
plot(CH4 ~ time, data = out2, type = 'l')
plot(VSd ~ time, data = out2, type = 'l')
plot(CH3COOH ~ time, data = out2, type = 'l')
plot(m1 ~ time, data = out2, type = 'l')
```

# 2. Time-variable inputs
Time-variable inputs (now separate from slurry mass series data) can be combined with both of the above examples.
First create the var_pars object.

```{r}
temp_series <- data.frame(
  time = c(0, 1, 10, 15, 30), 
  temp_C = c(10, 20, 25, 15, 10)
)
var_pars <- list(var = temp_series)
```

Use with series data.

```{r}
devtools::load_all()
out3 <- abm(
  structure = structure1,
  50,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  var_pars = var_pars
)
```

```{r}
plot(temp_C ~ time, data = out3, type = 'l')
plot(CH4_emis_rate ~ time, data = out3, type = 'l')
```

Or regular inputs.

```{r}
devtools::load_all()
out4 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  var_pars = var_pars
)
```

```{r}
plot(slurry_mass ~ time, data = out4, type = 'l')
plot(temp_C ~ time, data = out4, type = 'l')
plot(CH4_emis_rate ~ time, data = out4, type = 'l')
```

# 3. Flexible substrates

Above only a single particulate substrate was used.
But any, with arbitrary names, can be defined.

```{r}
sub_pars2 <- list(
  subs = c('slow', 'mid', 'fast'),
  T_opt_hyd = c(default = 60),
  T_min_hyd = c(default = 0),
  T_max_hyd = c(default = 90),
  hydrol_opt = c(slow = 0.01, mid = 0.1, fast = 1),
  sub_fresh = c(slow = 20, mid = 20, fast = 20),
  sub_init = c(slow = 20, mid = 20, fast = 20)
)
```
Q

```{r}
devtools::load_all()
out5 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars2,
  chem_pars = chem_pars
)
```

```{r}
plot(fast_conc ~ time, data = out5, type = 'l', ylim = c(0, 100))
lines(slow_conc ~ time, data = out5, col = 'red')
lines(CH3COOH_conc ~ time, data = out5, col = 'blue')

plot(fast ~ time, data = out5, type = 'l', ylim = c(0, 5e6))
lines(slow ~ time, data = out5, col = 'red')
lines(CH3COOH ~ time, data = out5, col = 'blue')
```

This is the simplest way to specify substrates.
When entered in this way, there is no stoichiometry *per se*, but each g substrate COD produces 1 g CH3COOH intermediate.

# 4. Sulfate reduction and other metabolic pathways
The microbial metabolism code now used is completely flexibile with respect to reactants and products.
Sulfate reduction is not hard-coded but can easily and very explicitly be specified with the new approach.
We need the appropriate stoichiometry.
Masses are specified as:

1. COD mass, or if COD = 0,
2. N mass, or if N = 0,
3. C mass, or if C = 0,
4. S mass, or if S = 0,
5. total mass

So for SO4-2, we want the the mass of S divided by the COD of acetate, i.e., 32 / 64 = 0.5.
(We might think about flexibility in how stoichiometry is defined, e.g., allow user to input moles, but it is probably better to fix it.)

```{r}
mstoich2 <- matrix(
  c(-1, -1, -1, -1,
     0,  0,  0, -0.5,
     1,  1,  1,  0),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'SO4', 'CH4'),
    c('m0', 'm1', 'm2', 'sr1')
  )
)

mstoich2

ksmat2 <- matrix(
  c(0.5, 0.5, 0.5, 0.5,
    NA,   NA,  NA, 0.5),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'SO4'),
    c('m0', 'm1', 'm2',  'sr1')
  )
)

ksmat2
```

Put these in a grp_pars list.

```{r}
grp_pars2 <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat2,
  mstoich = mstoich2
)
```

Sulfate needs to be added as a component.

```{r}
man_pars2 <- list(VFA_fresh = c(CH3COOH = 2), 
		  VFA_init = c(CH3COOH = 2),
		  comps = c('SO4'),
                  comp_fresh = c(SO4 = 10), 
                  comp_init = c(SO4 = 10), 
                  pH = 7, 
		  dens = 1000)
```

```{r}
devtools::load_all()
out6 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars2,
  grp_pars = grp_pars2,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
)
```

```{r}
plot(slurry_mass ~ time, data = out6, type = 'l')
plot(temp_C ~ time, data = out6, type = 'l')
plot(CH4_emis_rate ~ time, data = out4, type = 'l')
lines(CH4_emis_rate ~ time, data = out6, type = 'l', col = 'red')
plot(CH4 ~ time, data = out6, type = 'l')
plot(SO4_conc ~ time, data = out6, type = 'l')
plot(sr1 ~ time, data = out6, type = 'l')
plot(m1 ~ time, data = out6, type = 'l')
plot(CH3COOH_conc ~ time, data = out4, type = 'l')
lines(CH3COOH_conc ~ time, data = out6, type = 'l', col = 'red')
```

Notice that the sulfate reduction reaction is not balanced.
Neither is methanogenesis, for that matter.
Additional products can be added pretty easily (and would be included in default parameter sets presumably).

Let's add H2S and CO2.
Note that the coefficient on CO2 is g C, and g S on H2S and SO4.

```{r}
mstoich3 <- matrix(
  c(-1,  -1,  -1,  -1,
     0,   0,   0,  -0.5,
     1.5, 1.5, 1.5, 3,
     0,   0,   0,   0.5,
     1,   1,   1,   0),
  nrow = 5,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'SO4', 'CO2', 'H2S', 'CH4'),
    c('m0', 'm1', 'm2', 'sr1')
  )
)

mstoich3
```

```{r}
grp_pars3 <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat2,
  mstoich = mstoich3
)
```

Now we need more components.

```{r}
man_pars3 <- list(VFA_fresh = c(CH3COOH = 2), 
		  VFA_init = c(CH3COOH = 2),
		  comps = c('SO4', 'CO2', 'H2S'),
                  comp_fresh = c(SO4 = 10, CO2 = 1, H2S = 0), 
                  comp_init = c(SO4 = 10, CO2 = 1, H2S = 0), 
                  pH = 7, 
		  dens = 1000)
```

And we should specify that H2S and CO2 are gases, otherwise they accumulate like solutes.

```{r}
chem_pars2 <- list(COD_conv = c(CH4 = 3.01), gases = c('CH4', 'CO2', 'H2S'))
```

Remember that all this could be added to default parameter sets!

```{r}
devtools::load_all()
out7 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
)
```

```{r}
plot(CO2 ~ time, data = out7, type = 'l', col = 'blue')
lines(CH4 ~ time, data = out7, type = 'l')
lines(CH4 ~ time, data = out6, type = 'l', col = 'orange', lty = 2)
lines(H2S ~ time, data = out7, type = 'l', col = 'yellow')
plot(SO4_conc ~ time, data = out7, type = 'l')
plot(sr1 ~ time, data = out7, type = 'l')
plot(m1 ~ time, data = out7, type = 'l')
plot(CH3COOH_conc ~ time, data = out4, type = 'l')
lines(CH3COOH_conc ~ time, data = out7, type = 'l', col = 'red')
```

Repeat without any sulfate for sulfate reducers.

```{r}
man_pars4 <- list(VFA_fresh = c(CH3COOH = 2), 
		  VFA_init = c(CH3COOH = 2),
		  comps = c('SO4', 'CO2', 'H2S'),
                  comp_fresh = c(SO4 = 0, CO2 = 1, H2S = 0), 
                  comp_init = c(SO4 = 0, CO2 = 1, H2S = 0), 
                  pH = 7, 
		  dens = 1000)
```


```{r}
devtools::load_all()
out8 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars4,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
)
```
Q

```{r}
plot(CH4 ~ time, data = out8, type = 'l')
lines(CO2 ~ time, data = out8, type = 'l', col = 'blue')
lines(CH4 ~ time, data = out6, type = 'l', col = 'orange', lty = 2)
lines(H2S ~ time, data = out8, type = 'l', col = 'yellow')
```


Stoichiometry is all sorted out based on the chemical formula in `forms`, including how much VFA (if any) is produced and whether there is N mineralization (but then TAN needs to be added as a solute).




#mass_structure = list(
#  type = 'series',
#  data = 
#
#slurry_prod_rate = 10000, 
#                 slurry_mass = 1000,     
#                 resid_depth = 0.1,      
#                 empty_int = 100,          
#                 wash_water = 0,            
#                 wash_int = NA,
#                 rest_d = 0,


# Note
The "late" and "mid" options are still available, but now through `ctrl_pars`.
Here we can change the value through `add_pars`

# 5. Time-variable inputs part 2
Almost any model parameter can now be varied over time through var_pars.
If it is a single value without named elements for any time, it can be combined in the var data frame.
This excludes anything having to do with emptying, loading rate, or slurry level--if any of those need to vary non-regularly over time, the structure argument must be used!
So, that leaves probably only pH and temperature.
For example, we could simulate acidification with this.

```{r}
var_dat <- data.frame(
  time = c(0, 5, 10, 15, 30), 
  temp_C = c(10, 20, 25, 15, 10),
  pH = c(7, 6.5, 5, 5, 5)
)
var_pars2 <- list(var = var_dat)
```

```{r}
devtools::load_all()
out9 <- abm(
  50,
  structure = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
  var_pars = var_pars2
)
```

```{r}
plot(slurry_mass ~ time, data = out9, type = 'l')
plot(temp_C ~ time, data = out9, type = 'l')
plot(pH ~ time, data = out9, type = 'l')
plot(CH4 ~ time, data = out9, type = 'l')
lines(H2S ~ time, data = out9, type = 'l', col = 'brown')
```

For other inputs that are set in named elements, make a separate data frame for each other argument with column names that give the element names.
For example, for changing substrate concentration,

```{r}
sub_fresh_dat = data.frame(time = c(0, 5, 10, 15, 30), 
                       VSd = c(50, 100, 0, 0, 200))
```

and a sudden increase in sulfate, 

```{r}
comp_fresh_dat = data.frame(time = c(0, 5, 10, 15, 30), 
                       SO4 = c(0, 0, 0, 5, 5))
```

or even a change in microbial parameters,

```{r}
qhat_opt_dat = data.frame(time = c(0, 5, 10, 15, 30), 
                       m0 = c(1, 1, 1, 1, 0.1),
                       m1 = c(1, 1, 1, 1, 0.1),
                       m2 = c(1, 1, 1, 1, 0.1),
                       sr1 = c(2, 2, 2, 1, 1))
```

Then combine them all in a list, using the parameter element names for element names (e.g., `sub_fresh` is the name of an element in `sub_pars`).

```{r}
var_pars3 <- list(var = var_dat, sub_fresh = sub_fresh_dat, comp_fresh = comp_fresh_dat, qhat_opt = qhat_opt_dat)
```

Presently times need to be exactly aligned among all the var_par elements, and I might keep this behavior.

```{r}
devtools::load_all()
out10 <- abm(
  structure = structure2,
  50,
  mng_pars = mng_pars,
  man_pars = man_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
  var_pars = var_pars3
)
```

```{r}
plot(slurry_mass ~ time, data = out10, type = 'l')
plot(CH4_emis_rate ~ time, data = out10, type = 'l')
plot(VSd_conc ~ time, data = out10, type = 'l')
plot(SO4_conc ~ time, data = out10, type = 'l')
plot(CH3COOH_conc ~ time, data = out5, type = 'l')
```

Incredible flexibility, I think.

# 6. Flexible solutes and sulfate reduction
Any conservative solute can be added in `man_pars`, using any names.
I am moving toward using a "master species" approach, so it makes sense to use the chemical formula of the primary species, with `p` or `m` for a charge symbol.
The `comp` part of the name below is for "component".

```{r}
man_pars6 <- list(comps = c('H2S', 'SO4', 'NH4p'),
                 comp_fresh = c(H2S = 0.01, SO4 = 0.2, NH4p = 2.5), 
                 VFA_fresh = c(CH3COOH = 2),
                 pH = 7, dens = 1000)
```

Note that `CH3COOH` is still special--it has a fixed name in the code and is not conservative.


```{r}
devtools::load_all()
out6a <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars6,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars,
            chem_pars = chem_pars)
```

```{r}
tail(out6a)
head(out6a)
```

Conservative components are boring in output (without inhibition or volatilization).

```{r}
plot(NH4p_conc ~ time, data = out6a)
plot(NH4p ~ time, data = out6a)
```

But now with sulfate and sulfide, sulfate reducers can grow.

```{r}
plot(sr1_conc ~ time, data = out6a, type = 'l', ylim = c(0, 0.4))
lines(m1_conc ~ time, data = out6a, col = 'red')
```

And if H2S is a component, it will be produced.

```{r}
plot(SO4_conc ~ time, data = out6a, type = 'l')
plot(H2S_conc ~ time, data = out6a, type = 'l')
```

We can see dilution effects at if some washing water is added.

```{r}
mng_pars6 = list(slurry_prod_rate = 10000, 
                slurry_mass = 1000,     
                storage_depth = 2,     
                resid_depth = 0.1,      
                area = 100,              
                empty_int = 100,          
                temp_C = 20,
                wash_water = 100000,            
                wash_int = 100,
                rest_d = 0,
                resid_enrich = 1)

```

```{r}
out6b <- abm(365,
            mng_pars = mng_pars6,
            man_pars = man_pars6,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars,
            chem_pars = chem_pars)
```

```{r}
plot(NH4p_conc ~ time, data = out6b, type = 'l')
lines(NH4p_conc ~ time, data = out6a, col = 'red')
```

To include sulfate reduction, there must be an `sr` microbial group and `SO4` must be a chemical component.
Omit either to exclude sulfate reduction.

```{r}
man_pars6c <- list(comps = c('H2S', 'NH4p'),
                 comp_fresh = c(H2S = 0.01, NH4p = 2.5), 
                 VFA_fresh = c(CH3COOH = 2),
                 pH = 7, dens = 1000)
```

```{r}
grp_pars6c <- list(grps = c('m0', 'm1', 'm2'),
                 yield = c(default = 0.05),
                 xa_fresh = c(default = 0.05),
                 xa_init = c(default = 0.05),
                 dd_rate = c(default = 0.02),
                 ksv = c(default = 1),
                 qhat_opt = c(m0 = 1, m1 = 1, m2 = 2),
                 T_opt = c(m0 = 18, m1 = 18, m2 = 28),
                 T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41),
                 T_max = c(m0 = 25, m1 = 25, m2 = 38))
```

```{r}
devtools::load_all()
out6c <- abm(365,
            mng_pars = mng_pars6,
            man_pars = man_pars6c,
            grp_pars = grp_pars6c,
            mic_pars = mic_pars,
            sub_pars = sub_pars,
            chem_pars = chem_pars)
```

```{r}
plot(CH4_emis_cum ~ time, data = out6a, type = 'l')
lines(CH4_emis_cum ~ time, data = out6c, col = 'red')
```

```{r}
plot(CH3COOH_conc ~ time, data = out6a, type = 'l')
lines(CH3COOH_conc ~ time, data = out6c, col = 'red')
```

```{r}
plot(m1_conc ~ time, data = out6a, type = 'l')
lines(m1_conc ~ time, data = out6c, col = 'red')
```

```{r}
plot(slurry_mass ~ time, data = out6a, type = 'l')
lines(slurry_mass ~ time, data = out6c, col = 'red')
```

# 7. Speciation
Acid-base reactions are needed for inhibition and for CO2 emission.
They can be added for any component.
The `chem_pars` argument can accept temperature-dependent log ka expressions.
Use `temp_K` for absolute temperature in those expressions.

```{r}
man_pars7 <- list(comps = c('H2S', 'NH4p'),
                 comp_fresh = c(H2S = 0.01, NH4p = 2.5), 
                 VFA_fresh = c(CH3COOH = 2),
                 pH = 7, dens = 1000)

chem_pars7 <- list(COD_conv = c(CH4 = 1/0.2507, xa = 1/0.7069561,
                               VFA = 1/0.9383125, S = 1/0.5015, VS = 1/0.69, 
                               CO2_aer = 1/0.436, CO2_sr = 1/1.2, 
                               C_xa = 1/0.3753125),
                   specs = c('NH3', 'HSm', 'CH3COOm'),
                   mspec = c(NH3 = 'NH4p', HSm = 'H2S', CH3COOm = 'CH3COOH'),
                   lka = c(NH3 = '- 0.09046 - 2729.31/temp_K', 
                           HSm = '- 3448.7/temp_K + 47.479 - 7.5227 * log(temp_K)',
                           CH3COOm = '- 4.8288 + 21.42/temp_K')
)
```

Here `comps` are the chemical "components", or "master species", as described a bit above.
All are automatically included as chemical species in `packPars()`.
In the `chem_pars` argument, `specs` are the other (non-master) species that the master species are in equilirbium with.
And `mspec` are the associated master species.
So the NH3 species comes from NH4+.
When there is speciation, the master species are always taken as the protonated ones.
So the species in `specs` always have one less proton than their associated master species.
Only 2 species are supported for any component (master species and one more).

```{r}
devtools::load_all()
out7 <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars6,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars,
            chem_pars = chem_pars7)
```

```{r}
head(out7)
```

But chemical species don't matter unless they are used in inhibition or emission.

# 8. Inhibition

Any chemical species can inhibit any microbial group.
Inhibition parameters (currently initial and complete concentrations, with linear response, why not?) are entered in a matrix.

```{r}
ilwr <- matrix(
  c(5, 0.2, 10, 0.5,
    5, 0.2, 10, 0.5,
    5, 0.2, 10, 0.5,
    5, 0.2, 10, 0.5),
  nrow = 4,
  byrow = TRUE,
  dimnames = list(
    c('m0', 'm1', 'm2', 'sr1'),
    c('NH4p', 'NH3', 'CH3COOm', 'CH3COOH')
  )
)

iupr <- matrix(
  c(9, 0.9, 30, 1,
    9, 0.9, 30, 1,
    9, 0.9, 30, 1,
    9, 0.9, 30, 1),
  nrow = 4,
  byrow = TRUE,
  dimnames = list(
    c('m0', 'm1', 'm2', 'sr1'),
    c('NH4p', 'NH3', 'CH3COOm', 'CH3COOH')
  )
)

inhib_pars <- list(
  ilwr = ilwr,
  iupr = iupr
)

inhib_pars
                  
```

```{r}
man_pars8 <- list(comps = c('H2S', 'NH4p'),
                 comp_fresh = c(H2S = 0.01, 
				                        NH4p = 2.5), 
                 VFA_fresh = c(CH3COOH = 2),
                 pH = 7, dens = 1000)

chem_pars8 <- list(COD_conv = c(CH4 = 1/0.2507, xa = 1/0.7069561,
                               CH3COOH = 1/0.9383125, S = 1/0.5015, VS = 1/0.69, 
                               CO2_aer = 1/0.436, CO2_sr = 1/1.2, 
                               C_xa = 1/0.3753125),
                   specs = c('NH3', 'HSm', 'CH3COOm'),
                   mspec = c(NH3 = 'NH4p', HSm = 'H2S', CH3COOm = 'CH3COOH'),
                   lka = c(NH3 = '- 0.09046 - 2729.31/temp_K', 
                           HSm = '- 3448.7/temp_K + 47.479 - 7.5227* log(temp_K)',
                           CH3COOm = '-4.8288 + 21.42/temp_K')
)
```


```{r}
out8 <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars7,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars,
            chem_pars = chem_pars7,
            inhib_pars = inhib_pars
)
```

```{r}
plot(CH4_emis_rate ~ time, data = out7, type = 'l')
lines(CH4_emis_rate ~ time, data = out8, col = 'red')

plot(CH4_emis_cum ~ time, data = out7, type = 'l')
lines(CH4_emis_cum ~ time, data = out8, col = 'red')
```

# 9. Volatilization
Any chemical species can volatilize.

```{r}
man_pars9 <- list(comps = c('H2S', 'NH4p'),
                 comp_fresh = c(H2S = 0.01, NH4p = 2.5), 
                 VFA_fresh = c(CH3COOH = 2),
                 pH = 7, dens = 1000)
```

We need to set mass transfer coefficient values (m/d) for any species that volatilizes.
These are overall (possibly two-film) values for in liquid-phase units.

```{r}
chem_pars9 <- list(COD_conv = c(CH4 = 1/0.2507),
                   specs = c('NH3', 'HSm', 'CH3COOm'),
                   mspec = c(NH3 = 'NH4p', HSm = 'H2S', CH3COOm = 'CH3COOH'),
                   lka = c(NH3 = '- 0.09046 - 2729.31/temp_K', 
                           HSm = '- 3448.7/temp_K + 47.479 - 7.5227 * log(temp_K)',
                           CH3COOm = '-4.8288 + 21.42/temp_K'),
                   kl = c(NH3 = 0.01, H2S = 0.01, CH3COOH = 0.01) * 86400)

```


```{r}
devtools::load_all()
out9a <- abm(365,
             mng_pars = mng_pars,
             man_pars = man_pars9,
             grp_pars = grp_pars,
             mic_pars = mic_pars,
             sub_pars = sub_pars,
             chem_pars = chem_pars9,
             inhib_pars = inhib_pars
)
```

The state variable vector and output data frame automatically expand for the new values.

```{r}
head(out9a[, grep('_emis', names(out9a))])
```

```{r}
plot(NH4p_conc ~ time, data = out8, type = 'l', ylim = c(0, 3))
lines(NH4p_conc ~ time, data = out9a, col = 'red')
plot(NH4p ~ time, data = out8, type = 'l')
lines(NH4p ~ time, data = out9a, col = 'red')
plot(NH3_emis_cum ~ time, data = out9a, type = 'l')
plot(H2S_emis_cum ~ time, data = out9a, type = 'l')
plot(CH3COOH_emis_cum ~ time, data = out9a, type = 'l')
```

Let's use a fixed slurry mass to exaggerate emission.

```{r}
mng_pars9b = list(slurry_prod_rate = 0, 
                  slurry_mass = 1E6,     
                  storage_depth = 2,     
                  resid_depth = 0.1,      
                  area = 100,              
                  empty_int = 100,          
                  temp_C = 20,
                  wash_water = 0,            
                  wash_int = NA,
                  rest_d = 0,
                  resid_enrich = 1)
```

```{r}
chem_pars9b <- list(COD_conv = c(CH4 = 1/0.2507, xa = 1/0.7069561,
                               VFA = 1/0.9383125, S = 1/0.5015, VS = 1/0.69, 
                               CO2_aer = 1/0.436, CO2_sr = 1/1.2, 
                               C_xa = 1/0.3753125),
                   specs = c('NH3', 'HSm', 'CH3COOm'),
                   mspec = c(NH3 = 'NH4p', HSm = 'H2S', CH3COOm = 'CH3COOH'),
                   lka = c(NH3 = '- 0.09046 - 2729.31/temp_K', 
                           HSm = '- 3448.7/temp_K + 47.479 - 7.5227 * log(temp_K)',
                           CH3COOm = '-4.8288 + 21.42/temp_K'),
                   kl = c(NH3 = 0.01, H2S = 0.01) * 86400)

```



```{r}
devtools::load_all()
out9b <- abm(365,
             mng_pars = mng_pars9b,
             man_pars = man_pars9,
             grp_pars = grp_pars,
             mic_pars = mic_pars,
             sub_pars = sub_pars,
             chem_pars = chem_pars9b,
             inhib_pars = inhib_pars
)
```

```{r}
plot(NH4p ~ time, data = out9b, type = 'l')
plot(NH4p_conc ~ time, data = out9b, type = 'l')
```

# 9. COD balance
There is now a `checkCOD()` function that runs on `abm()` results before returning them.
For now the tolerance is fixed at 1%.
Some of the examples above do not meet that criterion for some reason.
At least one shows a real problem that needs to be identified.
For the emission example above, the problem is that VFA is emitted but that loss is not included in the balance check (this might have been fixed).
I need to decide about how to pass that COD information.
We can make it worse by pretending the the charged form can volatilize (`VFA` changed to `CH3COOm` below).
 
```{r}
chem_pars10 <- list(COD_conv = c(CH4 = 1/0.2507, xa = 1/0.7069561,
                               VFA = 1/0.9383125, S = 1/0.5015, VS = 1/0.69, 
                               CO2_aer = 1/0.436, CO2_sr = 1/1.2, 
                               C_xa = 1/0.3753125),
                   specs = c('NH3', 'HSm', 'CH3COOm'),
                   mspec = c(NH3 = 'NH4p', HSm = 'H2S', CH3COOm = 'CH3COOH'),
                   lka = c(NH3 = '- 0.09046 - 2729.31/temp_K', 
                           HSm = '- 3448.7/temp_K + 47.479 - 7.5227 * log(temp_K)',
                           CH3COOm = '-4.8288 + 21.42/temp_K'),
                   kl = c(NH3 = 0.01, H2S = 0.01, CH3COOm = 0.01) * 86400)

```

```{r}
out10 <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars9,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars,
            chem_pars = chem_pars10,
            inhib_pars = inhib_pars
)
```


# 10. Stoichiometry and nitrogen mineralization

Now substrates can produce any amount of arbitrary components (defined in `man_pars`, possibly volatilized, possibly involved in speciation in inhibition) through hydrolysis and fermentation to VFA.

```{r}
man_pars10 <- list(comps = c('H2S', 'SO4', 'NH4p'),
                   comp_fresh = c(H2S = 0.01, SO4 = 0.2, NH4p = 2.5), 
                   VFA_fresh = c(CH3COOH = 2),
                   pH = 7, dens = 1000)
```

(Hmm, should `comps` be moved to `chem_pars`?)

Here we'll have 4 substrates.
But substrates need not actually produce VFA anymore.

```{r}
sub_pars10 <- list(subs = c('cellulose', 'protein', 'lipids', 'urea'),
                   T_opt_hyd = c(default = 60),
                   T_min_hyd = c(default = 0),
                   T_max_hyd = c(default = 90),
                   hydrol_opt = c(lipids = 0.1, protein = 0.01, cellulose = 0.05, urea = 1),
                   sub_fresh = c(lipids = 3, protein = 20, cellulose = 35, urea = 10),
                   sub_init = c(lipids = 3, protein = 20, cellulose = 35, urea = 10))
```

Production of any component is set in the `fstoich` element of the `chem_pars` argument.


```{r}
smat <- matrix(c(0, 0.2,  0, 0.2,
                 0, 0.01, 0, 0,
                 1, 1,    1, 0),
               nrow = 3,
               byrow = TRUE,
               dimnames = list(
                 c('NH4p', 'H2S', 'CH3COOH'),
                 c('cellulose', 'protein', 'lipids', 'urea')))

smat
```

Substrate and other component quantities are

1. COD mass, or if COD = 0,
2. N mass, or if N = 0,
3. C mass, or if C = 0,
4. S mass, or if S = 0,
5. total mass

So the `CH3COOH` row should only have 1 or 0.

The `fstoich` matrix can be calculated from substrate chemical formulas--see next example.
               
```{r}
chem_pars10 <- list(COD_conv = c(CH4 = 1/0.2507, xa = 1/0.7069561,
                               VFA = 1/0.9383125, S = 1/0.5015, VS = 1/0.69, 
                               CO2_aer = 1/0.436, CO2_sr = 1/1.2, 
                               C_xa = 1/0.3753125),
                    specs = c('NH3', 'HSm', 'CH3COOm'),
                    mspec = c(NH3 = 'NH4p', HSm = 'H2S', CH3COOm = 'CH3COOH'),
                    fstoich = smat)

```

```{r}
devtools::load_all()
out10 <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars10,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars10,
            chem_pars = chem_pars10)
```

```{r}
plot(CH3COOH_conc ~ time, data = out10, type = 'l')
plot(urea_conc ~ time, data = out10, type = 'l', ylab = 'Urea or TAN (g/kg)')
lines(NH4p_conc ~ time, data = out10, col = 'blue')
```

It might make more sense to have `fstoich` calculated internally.
For this, substrate formulas must be provided.

```{r}
sub_pars10b <- list(subs = c('cellulose', 'protein', 'lipids', 'urea'),
                    forms = c(cellulose = 'C6H10O5', protein = 'C4 H6.1 O1.2 N',
                              lipids = 'C57 H104 O6', urea = 'CO(NH2)2'),
                    T_opt_hyd = c(default = 60),
                    T_min_hyd = c(default = 0),
                    T_max_hyd = c(default = 90),
                    hydrol_opt = c(lipids = 0.1, protein = 0.01, cellulose = 0.05, urea = 1),
                    sub_fresh = c(lipids = 3, protein = 20, cellulose = 35, urea = 10),
                    sub_init = c(lipids = 3, protein = 20, cellulose = 35, urea = 10))
```

Internally, the `getStoich()` function is used, which in turn calls `predFerm()`.
Among these substrates, cellulose produces no CO2 from fermentation, protein and lipids *consume* CO2 (they are highly reduced), and urea is a special case with no COD. 
So it produces no VFAs.

```{r}
predFerm(sub_pars10b$forms[1])
predFerm(sub_pars10b$forms[2])
predFerm(sub_pars10b$forms[3])
predFerm(sub_pars10b$forms[4])
```

Internally, the stiochiometric coefficients are changed to the same units given above (COD, N, C, S, or total mass).
Now we must have CO2 as a component, because it will be produced.

```{r}
man_pars10b <- list(comps = c('H2S', 'SO4', 'NH4p', 'CO2'),
                   comp_fresh = c(H2S = 0.01, SO4 = 0.2, NH4p = 2.5, CO2 = 1), 
                   VFA_fresh = c(CH3COOH = 2),
                   pH = 7, dens = 1000)
```

We don't actually need the species included above.
But because of a bit of a programming quirk, we need to provide the master species for NH3.

               
```{r}
chem_pars10b <- list(COD_conv = c(CH4 = 1/0.2507),
                    specs = c('NH3'),
                    mspec = c(NH3 = 'NH4p'),
                    fstoich = 'calc')

```

```{r}
devtools::load_all()
out10b <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars10b,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars10b,
            chem_pars = chem_pars10b)
```

```{r}
plot(CH3COOH_conc ~ time, data = out10b, type = 'l')
plot(urea_conc ~ time, data = out10b, type = 'l', ylab = 'Urea or TAN (g/kg)')
lines(NH4p_conc ~ time, data = out10b, col = 'blue')
```

Now we also have dissolved CO2 (really TIC) concentration.
Only there is no emission, so it does not mean much.

```{r}
plot(CO2_conc ~ time, data = out10b, type = 'l')
```

# 11. CO2 emission

CO2 emission can be includes through the volatilization route.
It can be produced from both fermentation and methanogenesis.
Speciation of dissolved CO2 (really H2CO3*) and HCO3- needs to be included (although a reduced mass transfer coefficient could achieve the same effect).
So far, CO2 emission behavior is troublesome--it is difficult to produce plausible results.

```{r}
sub_pars11 <- list(subs = c('VSd'),
                   forms = c(VSd = 'C23H37O14N (CO2)1.5'),
                   T_opt_hyd = c(default = 60),
                   T_min_hyd = c(default = 0),
                   T_max_hyd = c(default = 90),
                   hydrol_opt = c(default = 0.1),
                   sub_fresh = c(VSd = 20),
                   sub_init = c(VSd = 20))

```

```{r}
man_pars11 <- list(comps = c('CO2', 'NH4p'),
                 comp_fresh = c(CO2 = 1, NH4p = 2.5), 
                 VFA_fresh = c(CH3COOH = 2),
                 pH = 7, dens = 1000)
```

Here we will define speciation only for TIC.
There is no way to include CO3-2 as a third species.

```{r}
chem_pars11 <- list(COD_conv = c(CH4 = 1/0.2507),
                     specs = c('HCO3m'),
                     mspec = c(HCO3m = 'CO2', NH3 = 'NH4p'),
                     lka = c(HCO3m = '-2.778 + -353.5305 -0.06092*temp_K + 21834.37/temp_K + 126.8339*log10(temp_K) -1684915/temp_K^2'),
                     fstoich = 'calc',
                     kl = c(CO2 = 0.01) * 86400)
```

```{r}
devtools::load_all()
out11 <- abm(365,
            mng_pars = mng_pars,
            man_pars = man_pars11,
            grp_pars = grp_pars,
            mic_pars = mic_pars,
            sub_pars = sub_pars11,
            chem_pars = chem_pars11)
```

TIC in slurry shows a lot of change over time, but can be kept away from 0 by adjusting kl.
The low slurry level leads to a high relative loss.

```{r}
plot(CO2_conc ~ time, data = out11, type = 'l', ylim = c(0, 3))
```

But now we can predict a CO2:CH4 ratio.
This is a cumulative value.

```{r}
plot(CO2_emis_cum / CH4_emis_cum ~ time, data = out11, type = 'l', ylab = 'CO2:CH4')
plot(CO2_emis_cum / CH4_emis_cum ~ time, data = out11, type = 'l', ylab = 'CO2:CH4', ylim = c(0, 5))
abline(h = 1, lty = 2)
```

Let's try some different substrates.

# 12. Respiration

To include respiration just add a `kl` value for `O2`.

```{r}
chem_pars12a <- list(COD_conv = c(CH4 = 1/0.2507))

chem_pars12b <- list(COD_conv = c(CH4 = 1/0.2507),
                     kl = c(O2 = 0.1) * 86400)

```


```{r}
devtools::load_all()
out12a <- abm(365,
             mng_pars = mng_pars,
             man_pars = man_pars,
             grp_pars = grp_pars,
             mic_pars = mic_pars,
             sub_pars = sub_pars,
             chem_pars = chem_pars12a,
)
```

```{r}
devtools::load_all()
out12b <- abm(365,
             mng_pars = mng_pars,
             man_pars = man_pars,
             grp_pars = grp_pars,
             mic_pars = mic_pars,
             sub_pars = sub_pars,
             chem_pars = chem_pars12b,
)
```

```{r}
plot(CH4_emis_cum ~ time, data = out12a, type = 'l')
lines(CH4_emis_cum ~ time, data = out12b, col = 'red')
```

To track CO2 emission, we need all the inputs used in example 11.


```{r}
man_pars12 <- list(comps = c('CO2', 'NH4p'),
                   comp_fresh = c(CO2 = 1, NH4p = 3), 
                   VFA_fresh = c(CH3COOH = 2),
                   pH = 7, dens = 1000)
```

```{r}
chem_pars12c <- list(COD_conv = c(CH4 = 1/0.2507),
                     specs = c('HCO3m'),
                     mspec = c(HCO3m = 'CO2', NH3 = 'NH4p'),
                     lka = c(HCO3m = '-2.778 + -353.5305 -0.06092*temp_K + 21834.37/temp_K + 126.8339*log10(temp_K) -1684915/temp_K^2'),
                     fstoich = 'calc',
                     kl = c(CO2 = 0.01) * 86400)
```

```{r}
chem_pars12d <- list(COD_conv = c(CH4 = 1/0.2507),
                     specs = c('HCO3m'),
                     mspec = c(HCO3m = 'CO2', NH3 = 'NH4p'),
                     lka = c(HCO3m = '-2.778 + -353.5305 -0.06092*temp_K + 21834.37/temp_K + 126.8339*log10(temp_K) -1684915/temp_K^2'),
                     fstoich = 'calc',
                     kl = c(O2 = 0.1, CO2 = 0.01) * 86400)
```

(Hmm, shouldn't `fstoich` be part of `sub_pars`?)

We also need stoichiometry for CO2, so we need a substrate chemical formula.

```{r}
sub_pars12 <- list(subs = c('VSd'),
                   forms = c(VSd = 'C23H37O14N (CO2)1.5'),
                   T_opt_hyd = c(default = 60),
                   T_min_hyd = c(default = 0),
                   T_max_hyd = c(default = 90),
                   hydrol_opt = c(default = 0.1),
                   sub_fresh = c(VSd = 20),
                   sub_init = c(VSd = 20))

```

```{r}
devtools::load_all()
out12c <- abm(365,
             mng_pars = mng_pars,
             man_pars = man_pars12,
             grp_pars = grp_pars,
             mic_pars = mic_pars,
             sub_pars = sub_pars12,
             chem_pars = chem_pars12c,
)
```

```{r}
devtools::load_all()
out12d <- abm(365,
             mng_pars = mng_pars,
             man_pars = man_pars12,
             grp_pars = grp_pars,
             mic_pars = mic_pars,
             sub_pars = sub_pars12,
             chem_pars = chem_pars12d,
)
```

See how COD balance is worse?


```{r}
plot(CH4_emis_cum ~ time, data = out12c, type = 'l')
lines(CH4_emis_cum ~ time, data = out12d, col = 'red')
```


```{r}
plot(CO2_emis_cum ~ time, data = out12c, type = 'l')
lines(CO2_emis_cum ~ time, data = out12d, col = 'red')
```


