---
title: 'Comparison of abmneo and ABM'
output: pdf_document
author: Sasha D. Hafner
date: "`r format(Sys.time(), '%d %B, %Y %H:%M')`"
---

# Overview
This demo is meant to compare and align abmneo with the ABM package.

```{r}
devtools::load_all()
```

Set pars

Original ABM values from `ABM_fit/functions/setInitPars.R` at 1f98ed2590948562c24df9ac7272bf9c85c82f34.

```{r, eval = FALSE}
grp_pars <- list(
  grps = c('m0', 'm1', 'm2','m3','m4','m5','sr1'),
  yield = c(default = 0.05, sr1 = 0.065),
  xa_fresh = c(default = 0.06, sr1 = 0.01),
  xa_init = c(all = 0.06, sr1 = 0.01),
  decay_rate = c(all = 0.02),
  ks_coefficient = c(default = 1.153337, sr1 = 0.461335),
  qhat_opt =  c(m0 = 0.46289, m1 = 0.74568, m2 = 0.9006605, m3 = 2.79672, m4 = 4.66012, m5 = 7.4601, sr1 = 8.3451),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, m3 = 36, m4 = 43.75, m5 = 55, sr1 = 43.75),
  T_min = c(m0 = 0, m1 = 9.67, m2 = 9.67, m3 = 15, m4 = 26.25, m5 = 30, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, m3 = 45, m4 = 51.25, m5 = 60, sr1 = 51.25),
  ki_NH3_min = c(all = 0.015),
  ki_NH3_max = c(all = 5.37),
  ki_NH4_min = c(all = 2.7),
  ki_NH4_max = c(all = 6.8492),
  ki_H2S_slope = c(default = -0.10623, sr1 = -0.1495),
  ki_H2S_int = c(default = 1.02, sr1 = 1.42),
  ki_H2S_min = c(default = 0.08),
  IC50_low = c(default = 0.20854, sr1 = 0.2772),
  ki_HAC = c(default = 0.31),
  pH_UL = c(default = 8.0),
  pH_LL = c(default = 6.5, sr1 = 5.5)
)
```

```{r}
mstoich <- matrix(
  c(
    -1, -1, -1, -1, -1, -1, -1,
     0,  0,  0,  0,  0,  0, -0.50156,
     1,  1,  1,  1,  1,  1,  0
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4', 'CH4'),
    c('m0', 'm1', 'm2', 'm3', 'm4', 'm5', 'sr1')
  )
)

mstoich

# SO4 is from mic.pars2.0 in ABM

ksmat <- matrix(
  c(1.15, 1.15, 1.15, 1.15, 1.15, 1.15, 0.46,
      NA,   NA,   NA,   NA,   NA,   NA, 0.00694),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4'),
    c('m0', 'm1', 'm2', 'm3', 'm4', 'm5', 'sr1')
  )
)

ksmat
```


```{r, eval = FALSE}
grp_ref <- list(
  grps = c('m0', 'm1', 'm2'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  qhat_opt = c(m0 = 0.5, m1 = 1, m2 = 1),
  T_opt = c(m0 = 10, m1 = 18, m2 = 28),
  T_min = c(m0 = 0, m1 = 5, m2 = 10),
  T_max = c(m0 = 25, m1 = 25, m2 = 38),
  ksmat = ksmat_ref,
  mstoich = mstoich_ref
)
```

```{r}
grp_pars <- list(
  grps = c('m0', 'm1', 'm2','m3','m4','m5','sr1'),
  yield = c(default = 0.05, sr1 = 0.065),
  xa_fresh = c(m0 = 0.06, m1 = 0.06, m2 = 0.06, m3 = 0.06, m4 = 0.06, m5 = 0.06, sr1 = 0.06),
  xa_init = c(default = 0.06),
  dd_rate = c(default = 0.02),
  qhat_opt =  c(m0 = 0.46289, m1 = 0.74568, m2 = 0.9006605, m3 = 2.79672, m4 = 4.66012, m5 = 7.4601, sr1 = 8),
  #qhat_opt =  c(default = 0, sr1 = 8.3451),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, m3 = 36, m4 = 43.75, m5 = 55, sr1 = 43.75),
  T_min = c(m0 = 0, m1 = 9.67, m2 = 9.67, m3 = 15, m4 = 26.25, m5 = 30, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, m3 = 45, m4 = 51.25, m5 = 60, sr1 = 51.25),
  ksmat = ksmat,
  mstoich = mstoich
)

grp_pars
```

Influent pars

```{r}
inf_pars <- list(
  VFA_fresh = 0, 
  VFA_init = 10,
  comps = c('SO4'),
  comp_fresh = c(SO4 = 1), 
  comp_init = c(SO4 = 1), 
  pH = 7, 
  dens = 1000
)
```

```{r}
mass_series <- read.csv('level_pig_od_ref.csv') 
plot(slurry_mass ~ time, data = mass_series, type = 'o')

stor_ser <- list(
  type = 'series',
  dat = mass_series,
  storage_depth = 2,     
  area = 0.7,              
  temp_C = 18,
  resid_enrich = 0.9 
)
```

Substrates and hydrolysis

```{r, eval = FALSE}
arrh_pars <- list(lnA =  29.96298,
              E_CH4 = 81000, 
              A = c(xa_dead= 3.61383 * 10^12, starch = 5 * 3.61383 * 10^12, Cfat = 0.1 * 3.61383 * 10^12, CPs = 3.61383 * 10^12, CPf = 3.61383 * 10^12, RFd = 3.61383 * 10^12, VSd = 3.61383 * 10^12, urea = 4.38*10^15), 
              E = c(xa_dead= 81557, starch = 81557, Cfat = 81557, CPs = 81557, CPf = 81557, RFd = 81557, VSd = 81557, urea = 81559),  
              R = 8.314,  
              VS_CH4 = 6.67,
              scale_alpha_opt = list(VSd = 0.46056, notVSd = 0.556886, CP = 2.065),
              kl = c(NH3 = 336, NH3_floor = 0.36*337, H2S = 0.02),
              scale_EF_NH3 = 0.0000048)
```

```{r}
Arrh_func <- function(A, E, R, temp_K){
  
  y <- A * exp(-E/(R * temp_K))
  
  return(y)
}
```

```{r}
r <- Arrh_func(A = 3.61383e12, E = 81557, R = 8.314, temp_K = (tempC <- 0:55) + 273.15)
args(abmneo:::CTM)
ctmv <- Vectorize(abmneo:::CTM)
rc <- ctmv(temp_K = (tempC <- 0:55) + 273.15, t_opt = 80 + 273.15, t_min = 0 + 273.15, t_max = 80 + 273.15, y_opt = 0.7)

plot(tempC, r, ylim = c(0, 1))
lines(tempC, rc)
```

```{r}
sub_pars <- list(
  subs = c('starch', 'Cfat', 'CPs', 'CPf', 'RFd', 'xd'),
  xd = 'xd',
  T_opt_hyd = c(default = 80),
  T_min_hyd = c(default = 0),
  T_max_hyd = c(default = 80),
  hydrol_opt = c(default = 0.7),
  sub_fresh = c(starch = 10, default = 0),
  sub_init = c(starch = 10, default = 0)
)
```

```{r}
ls()
```

Chem pars.
COD_conv is g COD per g whatever (here g CH4-C and g SO4-S)

```{r}
chem_pars <- list(COD_conv = c(CH4 = 5.32, SO4 = -1.9938), gases = c('CH4'))
```

Get COD of SO4-2
micstoich::micstoich('CH3COOH', acceptor = 'SO4-2', product = 'H2S')
How much COD consumed?
micstoich::calcCOD('CH3COOH', per = 'mol') * 0.125
How much SO4-S consumed?
micstoich::molmass('SO4-2', 'S') * 0.125
4.0125 / 8
8 / 4.0125
Try to run


Q
```{r}
devtools::load_all()
out <- abmneo(
  storage = stor_ser,
  365,
  inf_pars = inf_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 0
)
```
traceback()

```{r}
plot(slurry_mass ~ time, data = out, type = 'l')
plot(CH4_emis_rate ~ time, data = out, type = 'l')
plot(CH4 ~ time, data = out, type = 'l')
matplot(out$time, out[, nn <- c('m0','m1','m2', 'm3', 'm4', 'm5', 'sr1')], col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
matplot(out$time, out[, nn <- paste0(c('m0','m1','m2', 'm3', 'm4', 'm5', 'sr1'), '_conc')], col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
matplot(out$time, out[, nn <- paste0(c('starch', 'Cfat', 'CPs', 'CPf', 'RFd'), '_conc')], col = cc <- c('blue', 'orange', 'green', 'black', 'red'), lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
plot(VFA ~ time, data = out, type = 'l')
plot(starch_conc ~ time, data = out, type = 'l')
plot(xd_conc ~ time, data = out, type = 'l')
plot(VFA_conc ~ time, data = out, type = 'l')
plot(SO4_conc ~ time, data = out, type = 'l')
plot(COD_load ~ time, data = out, type = 'l')
```


