---
title: 'Demo of `abmneo()` function'
output: pdf_document
author: Sasha D. Hafner
date: "`r format(Sys.time(), '%d %B, %Y %H:%M')`"
---

# What's different?
The abmneo package was meant to address some issues with ABM:

1. complexity (some deliberate, some not),
2. hard-coded behavior and special cases,
3. number of dependencies,
4. lack of verification,
5. size of codebase, 
6. use of Rcpp, and
7. model speed.

On 1, the flexibility of ABM and the number of processes was all deliberate.
These are not inherently bad things, but they did complicate the model and especially the code, making it difficult to understand, which *is* bad.
Complexity also makes code verification, debugging, and maintenance difficult.
At some point in the future, as dependencies change (see point 3 as well), it can become too difficult.

Hard-coded behavior and special cases, such as the distinctions between methanogens and sulfate reducers, with their element names indicating what they were and which code should be used for them in `rates()` and probably elsewhere makes for complex, inefficient, and fragile code.
By fragile (a word I picked up from Claude) I mean likely to cause errors depending on inputs.
Inhibition and respiration are two other areas where there was hard-coded behavior and special cases.

Dependencies (3) is actually not much of an issue.
We had stats, dplyr (not a big fan), and of course Rcpp.
If we can avoid them, why not?
Rcpp is dicsussed below--that is significant, I think.

Verification is something that has really worried me.
ABM (and abmneo) is complex enough that mistakes in mass balance or elsewhere are quite likely.
So verification is important.

The size of the codebase (another term from Claude) alone also contributes to difficulty in understanding the package, and it seems like the possibility of holding most of even just the most important parts of the code in a developer's mind is impossible.
I gave up trying to follow the logic on more than one occassion.

Using Rcpp was of course done for speed, but it complicates the package with respect to debugging and development in general, and increases the requires for installation.
If it could be avoided without a speed penalty, wonderful.

Of course speed has always been an issue.
My hope was to not make it worse.

So, how is abmneo different?
I think I made significant progress on most of those points:

1. model and codebase are both much simpler,
2. eliminated most hard-coding; even microbial groups can have any metabolic pathways and any names,
3. deSolve is the only dependency,
4. COD balance is built in (could still be improved/extended),
5. size of the codebase is about half as large (2688 lines vs. 1393) and the `rates()` function in particular is shorter and clearer (563 vs. 50!),
6. Rcpp is no longer used, but
7. the `abmneo()` function *seems* quite fast.

Some of these changes are clear improvements (thinking about how microbial groups are handled now).
Many others mean fewer features and processes.
These are summarized in the demos below.

# Prep

```{r}
devtools::load_all()
```

# Some default parameters
First substrate parameters, a new argument compared to ABM().
This defines substrates.
We could have any number with any names but will just use the old `VSd` here.
Note that hydrolysis uses CTM again.

```{r}
sub_ref <- list(
  subs = c('VSd'),
  T_opt_hyd = c(VSd = 60),
  T_min_hyd = c(VSd = 0),
  T_max_hyd = c(VSd = 90),
  hydrol_opt = c(VSd = 0.1),
  sub_fresh = c(VSd = 30),
  sub_init = c(VSd = 30)
)
```

Microbial parameters are similar to other ABM versions.
But now we have microbial stoichiometry instead of the "special cases" that complicate the code.
The ks par is now in a matrix, which could have more than one row when there are multiple substrates in a reaction (e.g., with sulfate reduction).

```{r}
mstoich_ref <- matrix(
  c(-1, -1, -1, 1, 1, 1),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'CH4'),
    c('m0', 'm1', 'm2')
  )
)

mstoich_ref

ksmat_ref <- matrix(
  c(0.5, 0.5, 0.5),
  nrow = 1,
  byrow = TRUE,
  dimnames = list(
    c('VFA'),
    c('m0', 'm1', 'm2')
  )
)

ksmat_ref
```

With a COD basis for substrates and CH4 internally, stoichiometric coefficients are usually simple, as above.
These matrices are put into `grp_pars` along with recognizable elements.
Note that CTM is used for all termperature responses.

```{r}
grp_ref <- list(
  grps = c('m0', 'm1', 'm2'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  qhat_opt = c(m0 = 0.5, m1 = 1, m2 = 1),
  T_opt = c(m0 = 10, m1 = 18, m2 = 28),
  T_min = c(m0 = 0, m1 = 5, m2 = 10),
  T_max = c(m0 = 25, m1 = 25, m2 = 38),
  ksmat = ksmat_ref,
  mstoich = mstoich_ref
)
```

The `dd_rate` parameter is for "death and decay".

These last two parameter lists here are similar to other versions.
VFA is hard-coded as an intermediate.
The name `VFA` is used internally and in output.
No name is set here in the elements.
This `inf_pars` holds what is left of the original `man_pars`.
The name is for *inf* low, meant to be more generic than manure.

```{r}
inf_ref <- list(VFA_fresh = 2, VFA_init = 2, pH = 7, dens = 1000)
```

The chem_pars list now specifies which components are emitted as gases and which are otherwise solutes.
Gases have no association with the slurry like solutes or substrates do--their production is taken as emission and cumulative values are tracked.

```{r}
chem_ref <- list(COD_conv = c(CH4 = 5.32), gases = c('CH4'))
```

Conversion factor from: `calc_COD('CH4') / mol_mass('C')` from the `mic-stoich` repo.
It is g COD / g CH4-C.


# 1. Slurry mass series versus regular inputs
When it comes to how slurry level and the fill/empty schedule is entered, we now distinguish between *regular* inputs (as in repeating/consistent, as in earlier versions of `abmneo()`) and *series* inputs, where the schedule is inferred from time series data on slurry mass (what was called time-variable in earlier versions of `abmneo()`).
These are specified in the `storage` argument.
The name indicates that this argument describes the storage structure.
The slurry mass data frame for storage only has time and slurry_mass.
It is no longer a part of `var_pars`--the two are now independent (at least externally).

We'll work on a single pig scale here with some normal outdoor tank levels from a spreadsheet calculator from Peter Kai.

```{r}
mass_series <- read.csv('level_pig_od_ref.csv') 
plot(slurry_mass ~ time, data = mass_series, type = 'o')

stor_ser_ref <- list(
  type = 'series',
  dat = mass_series,
  storage_depth = 2,     
  area = 0.7,              
  temp_C = 18,
  resid_enrich = 0.2 
)
```

Unlike `ABM()`, no `var_pars` is needed.
And the fixed inputs like `slurry_prod_rate` that were ignored in these cases with `ABM()` are no longer expected.
(But the `storage` argument expects these when `type = 'regular'`.)

```{r}
outpodr <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  chem_pars = chem_ref
)
```

```{r}
plot(slurry_mass ~ time, data = outpodr, type = 'l')
plot(CH4_emis_rate ~ time, data = outpodr, type = 'l')
plot(CH4 ~ time, data = outpodr, type = 'l')
matplot(outpodr$time, outpodr[, nn <- c('m0','m1','m2')], col = 'black', lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, lty = 1:length(nn))
plot(VFA ~ time, data = outpodr, type = 'l')
plot(VFA_conc ~ time, data = outpodr, type = 'l')
plot(VSd_conc ~ time, data = outpodr, type = 'l')
plot(COD_load ~ time, data = outpodr, type = 'l')
```

With regular inputs, `slurry_prod_rate` etc. are now explicitly given.
This is also for one pig, with a 28 d batch period.

```{r}
stor_reg_ref <- list(
  type = 'regular',
  slurry_prod_rate = 5, 
  storage_depth = 2,     
  area = 0.65,              
  temp_C = 20,
  resid_enrich = 0.2,
  slurry_mass = 0.1,     
  resid_depth = 0.1,      
  empty_int = 28,          
  wash_water = 0,            
  wash_int = 90,
  rest_d = 1
)
```

```{r}
outpinr <- abmneo(
  storage = stor_reg_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  chem_pars = chem_ref
)
```

(Will get an error when starting with 0 slurry_mass)

```{r}
plot(slurry_mass ~ time, data = outpinr, type = 'l')
plot(CH4_emis_rate ~ time, data = outpinr, type = 'l')
plot(CH4 ~ time, data = outpinr, type = 'l')
matplot(outpinr$time, outpinr[, nn <- c('m0','m1','m2')], col = 'black', lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, lty = 1:length(nn))
plot(VFA ~ time, data = outpinr, type = 'l')
plot(VFA_conc ~ time, data = outpinr, type = 'l')
plot(VSd_conc ~ time, data = outpinr, type = 'l')
plot(COD_load ~ time, data = outpinr, type = 'l')
```

# 2. Time-variable inputs
Time-variable inputs (now separate from slurry mass series data) can be combined with both of the above examples.
Let's use it in the outdoor storage data (series type `storage`).

```{r}
temp_dat <- data.frame(time = seq(0, 365, length.out = 12))
temp_dat$temp_C <- 10 + 10 * sin((temp_dat$time - 100) / 365 * 2 * pi)
plot(temp_C ~ time, data = temp_dat, type = 'o')
```

```{r}
devtools::load_all()
outpodv <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  chem_pars = chem_ref,
  var_pars = list(var = temp_dat)
)
```

```{r}
plot(slurry_mass ~ time, data = outpodv, type = 'l')
plot(temp_C ~ time, data = outpodv, type = 'l')
plot(CH4_emis_rate ~ time, data = outpodr, type = 'l')
lines(CH4_emis_rate ~ time, data = outpodv, type = 'l', col = 'blue')
plot(CH4 ~ time, data = outpodr, type = 'l')
lines(CH4 ~ time, data = outpodv, type = 'l', col = 'blue')
matplot(outpodv$time, outpodv[, nn <- c('m0','m1','m2')], col = 'black', lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, lty = 1:length(nn))
plot(VFA_conc ~ time, data = outpodv, type = 'l')
plot(VSd_conc ~ time, data = outpodv, type = 'l')
```

Or regular inputs.

```{r}
outpinv <- abmneo(
  storage = stor_reg_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  chem_pars = chem_ref,
  var_pars = list(var = temp_dat)
)
```

```{r}
plot(slurry_mass ~ time, data = outpinv, type = 'l')
plot(temp_C ~ time, data = outpinv, type = 'l')
plot(CH4_emis_rate ~ time, data = outpinr, type = 'l')
lines(CH4_emis_rate ~ time, data = outpinv, type = 'l', col = 'blue')
plot(CH4 ~ time, data = outpinr, type = 'l')
lines(CH4 ~ time, data = outpinv, type = 'l', col = 'blue')
matplot(outpinv$time, outpinv[, nn <- c('m0','m1','m2')], col = 'black', lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, lty = 1:length(nn))
plot(VFA_conc ~ time, data = outpinv, type = 'l')
plot(VSd_conc ~ time, data = outpinv, type = 'l')
```

# 3. Flexible substrates

Above only a single particulate substrate was used.
But any substrates, with arbitrary names and stoichiometry can be defined.

```{r}
sub_smf <- list(
  subs = c('slow', 'mid', 'fast'),
  T_opt_hyd = c(default = 60),
  T_min_hyd = c(default = 0),
  T_max_hyd = c(default = 90),
  hydrol_opt = c(slow = 0.01, mid = 0.1, fast = 1),
  sub_fresh = c(slow = 10, mid = 10, fast = 10),
  sub_init = c(slow = 10, mid = 10, fast = 10)
)
```

```{r}
outpodf <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_smf,
  chem_pars = chem_ref
)
```

```{r}
plot(slurry_mass ~ time, data = outpodf, type = 'l')
plot(slow_conc ~ time, data = outpodf, type = 'l', ylim = c(0, 10))
lines(mid_conc ~ time, data = outpodf, col = 'blue')
lines(fast_conc ~ time, data = outpodf, col = 'red')
```

This is the simple way to specify substrates.
When entered in this way, there is no stoichiometry *per se*, but each g substrate COD produces 1 g VFA intermediate.

# 4. Sulfate reduction and other metabolic pathways
The microbial metabolism code (whatever happens after VFA intermediate is formed) now used is completely flexibile with respect to reactants and products.
Sulfate reduction is not hard-coded but can easily and very explicitly be specified with the new approach.
We need the appropriate stoichiometry.
Masses/coefficients are specified as:

1. COD mass, or if COD = 0, then
2. N mass, or if N = 0, then
3. C mass, or if C = 0, then
4. S mass, or if S = 0, then
5. total mass.

So for SO4-2, we want the the mass of S divided by the COD of acetate, i.e., 32 / 64 = 0.5.
(We might think about flexibility in how stoichiometry is defined, e.g., allow user to input moles, but it is probably better to fix it.)

```{r}
mstoich_sr <- matrix(
  c(-1, -1, -1, -1,
     0,  0,  0, -0.5,
     1,  1,  1,  0),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4', 'CH4'),
    c('m0', 'm1', 'm2', 'sr1')
  )
)

mstoich_sr

ksmat_sr <- matrix(
  c(0.5, 0.5, 0.5, 0.5,
    NA,   NA,  NA, 0.5),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4'),
    c('m0', 'm1', 'm2',  'sr1')
  )
)

ksmat_sr
```

Put these in a grp_pars list.

```{r}
grp_sr <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat_sr,
  mstoich = mstoich_sr
)
```

Sulfate needs to be added as a component.

```{r}
inf_sr <- list(VFA_fresh = 2, 
		  VFA_init = 2,
		  comps = c('SO4'),
                  comp_fresh = c(SO4 = 10), 
                  comp_init = c(SO4 = 10), 
                  pH = 7, 
		  dens = 1000)
```

Because it is a solute, we don't need to change the `chem_pars` contents (to indicated that it is a gas).

```{r}
outpodsr <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_sr,
  grp_pars = grp_sr,
  sub_pars = sub_ref,
  chem_pars = chem_ref
)
```

```{r}
plot(slurry_mass ~ time, data = outpodsr, type = 'l')
matplot(outpodsr$time, outpodsr[, nn <- c('m0','m1','m2', 'sr1')], col = 'black', lty = 1:length(nn), type = 'l')
legend('topright', legend = nn, lty = 1:length(nn))
plot(VSd_conc ~ time, data = outpodsr, type = 'l')
plot(VFA_conc ~ time, data = outpodsr, type = 'l')
plot(SO4_conc ~ time, data = outpodsr, type = 'l')
```

Notice that the sulfate reduction reaction is not balanced.
Neither is methanogenesis, for that matter.
Additional arbitrary chemical products can be added easily (and would be included in default parameter sets presumably).

Let's add H2S and CO2.
Note that the coefficient on CO2 is g C, and g S on H2S and SO4.

```{r}
#calc_COD('CH4')
#mol_mass('C') 
#12/64
```

```{r}
mstoich_src <- matrix(
  c(-1,    -1,    -1,    -1,
     0,     0,     0,    -0.5,
     0.187, 0.187, 0.187, 0.4,
     0,     0,     0,     0.5,
     1,     1,     1,     0),
  nrow = 5,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4', 'CO2', 'H2S', 'CH4'),
    c('m0', 'm1', 'm2', 'sr1')
  )
)

mstoich_src
```

Only that matrix needs to change.

```{r}
grp_pars_src <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat_sr,
  mstoich = mstoich_src
)
```

Now we need more components.

```{r}
inf_pars_src <- list(VFA_fresh =  2, 
		  VFA_init = 2,
		  comps = c('SO4', 'CO2', 'H2S'),
                  comp_fresh = c(SO4 = 10, CO2 = 1, H2S = 0), 
                  comp_init = c(SO4 = 10, CO2 = 1, H2S = 0), 
                  pH = 7, 
		  dens = 1000)
```

And we should specify that H2S and CO2 are gases, otherwise they accumulate like solutes.

```{r}
chem_pars_src <- list(COD_conv = c(CH4 = 5.32), gases = c('CH4', 'CO2', 'H2S'))
```

Remember that all this could be added to default parameter sets!

```{r}
devtools::load_all()
out7 <- abmneo(
  50,
  storage = storage2,
  inf_pars = inf_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
)
```

```{r}
plot(CO2 ~ time, data = out7, type = 'l', col = 'blue')
lines(CH4 ~ time, data = out7, type = 'l')
lines(CH4 ~ time, data = out6, type = 'l', col = 'orange', lty = 2)
lines(H2S ~ time, data = out7, type = 'l', col = 'yellow')
plot(SO4_conc ~ time, data = out7, type = 'l')
plot(sr1 ~ time, data = out7, type = 'l')
plot(m1 ~ time, data = out7, type = 'l')
plot(VFA_conc ~ time, data = out4, type = 'l')
lines(VFA_conc ~ time, data = out7, type = 'l', col = 'red')
```

Repeat without any sulfate for sulfate reducers.

```{r}
inf_pars4 <- list(VFA_fresh = 2, 
		  VFA_init = 2,
		  comps = c('SO4', 'CO2', 'H2S'),
                  comp_fresh = c(SO4 = 0, CO2 = 1, H2S = 0), 
                  comp_init = c(SO4 = 0, CO2 = 1, H2S = 0), 
                  pH = 7, 
		  dens = 1000)
```


```{r}
devtools::load_all()
out8 <- abmneo(
  50,
  storage = storage2,
  inf_pars = inf_pars4,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
)
```
Q

```{r}
plot(CH4 ~ time, data = out8, type = 'l')
lines(CO2 ~ time, data = out8, type = 'l', col = 'blue')
lines(CH4 ~ time, data = out6, type = 'l', col = 'orange', lty = 2)
lines(H2S ~ time, data = out8, type = 'l', col = 'yellow')
```


Stoichiometry is all sorted out based on the chemical formula in `forms`, including how much VFA (if any) is produced and whether there is N mineralization (but then TAN needs to be added as a solute).




#mass_storage = list(
#  type = 'series',
#  data = 
#
#slurry_prod_rate = 10000, 
#                 slurry_mass = 1000,     
#                 resid_depth = 0.1,      
#                 empty_int = 100,          
#                 wash_water = 0,            
#                 wash_int = NA,
#                 rest_d = 0,


# Note
The "late" and "mid" options are still available, but now through `ctrl_pars`.
Here we can change the value through `add_pars`

# 5. Time-variable inputs part 2
Almost any model parameter can now be varied over time through var_pars.
If it is a single value without named elements for any time, it can be combined in the var data frame.
This excludes anything having to do with emptying, loading rate, or slurry level--if any of those need to vary non-regularly over time, the storage argument must be used!
So, that leaves probably only pH and temperature.
For example, we could simulate acidification with this.

```{r}
var_dat <- data.frame(
  time = c(0, 5, 10, 15, 30), 
  temp_C = c(10, 20, 25, 15, 10),
  pH = c(7, 6.5, 5, 5, 5)
)
var_pars2 <- list(var = var_dat)
```

```{r}
devtools::load_all()
out9 <- abmneo(
  50,
  storage = storage2,
  inf_pars = inf_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
  var_pars = var_pars2
)
```

```{r}
plot(slurry_mass ~ time, data = out9, type = 'l')
plot(temp_C ~ time, data = out9, type = 'l')
plot(pH ~ time, data = out9, type = 'l')
plot(CH4 ~ time, data = out9, type = 'l')
lines(H2S ~ time, data = out9, type = 'l', col = 'brown')
```

For other inputs that are set in named elements, make a separate data frame for each other argument with column names that give the element names.
For example, for changing substrate concentration,

```{r}
sub_fresh_dat = data.frame(time = c(0, 5, 10, 15, 30), 
                       VSd = c(50, 100, 0, 0, 200))
```

and a sudden increase in sulfate, 

```{r}
comp_fresh_dat = data.frame(time = c(0, 15, 30), 
                       SO4 = c(0, 5, 5))
```

or even a change in microbial parameters,

```{r}
qhat_opt_dat = data.frame(
  time = c(0, 22), 
  m0   = c(1, 0.1),
  m1   = c(1, 0.1),
  m2   = c(1, 0.1),
  sr1  = c(2, 1)
)
```

Then combine them all in a list, using the parameter element names for element names (e.g., `sub_fresh` is the name of an element in `sub_pars`).

```{r}
var_pars3 <- list(var = var_dat, sub_fresh = sub_fresh_dat, comp_fresh = comp_fresh_dat, qhat_opt = qhat_opt_dat)
```

Note that times do not need to be the same!
The data frames are merged and missing values are interpolated or filled forward (we need to align the language for this and mid etc.)

```{r}
devtools::load_all()
out10 <- abmneo(
  storage = storage2,
  50,
  inf_pars = inf_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
  var_pars = var_pars3
)
```

```{r}
plot(slurry_mass ~ time, data = out10, type = 'l')
plot(CH4_emis_rate ~ time, data = out10, type = 'l')
plot(VSd_conc ~ time, data = out10, type = 'l')
plot(SO4_conc ~ time, data = out10, type = 'l')
plot(VFA_conc ~ time, data = out5, type = 'l')
```

That is a lot of flexibility, I think.

# X. The startup argument

Works as before.
This example is based on out2

```{r}
storage3 <- list(
  type = 'regular',
  slurry_prod_rate = 1E4, 
  storage_depth = 2,     
  area = 100,              
  temp_C = 20,
  resid_enrich = 0.2,
  slurry_mass = 5E4,     
  resid_depth = 0.5,      
  empty_int = 30,          
  wash_water = 100,            
  wash_int = NA,
  rest_d = 1
)
```

Reduce initial methanogen concentrations and set fresh to zero.

```{r}
grp_pars4 <- list(
  grps = c('m0', 'm1', 'm2'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.01),
  xa_init = c(default = 0.01),
  dd_rate = c(default = 0.02),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41),
  T_max = c(m0 = 25, m1 = 25, m2 = 38),
  ksmat = ksmat,
  mstoich = mstoich
)
```

```{r}
devtools::load_all()
out11_1 <- abmneo(
  90,
  delta_t = 0.1,
  storage = storage3,
  inf_pars = inf_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 1
)
```


```{r}
out11_0 <- abmneo(
  90,
  delta_t = 0.1,
  storage = storage3,
  inf_pars = inf_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 0
)

devtools::load_all()
out11_1 <- abmneo(
  90,
  delta_t = 0.1,
  storage = storage3,
  inf_pars = inf_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 1
)

out11_2 <- abmneo(
  90,
  delta_t = 0.1,
  storage = storage3,
  inf_pars = inf_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 2
)

out11_3 <- abmneo(
  90,
  delta_t = 0.1,
  storage = storage3,
  inf_pars = inf_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 10 
)

```

```{r}
out11_0$slurry_mass
out11_0$CH4_emis_rate
plot(slurry_mass ~ time, data = out11_0, type = 'l')

plot(CH4_emis_rate ~ time, data = out11_2, type = 'l', col = 'blue', ylim = c(0, 1.4E6))
lines(CH4_emis_rate ~ time, data = out11_1, col = 'red')
lines(CH4_emis_rate ~ time, data = out11_3, col = 'orange')
lines(CH4_emis_rate ~ time, data = out11_0, col = 'black')

plot(m1_conc ~ time, data = out11_2, type = 'l', col = 'blue', ylim = c(0, 0.06))
lines(m1_conc ~ time, data = out11_1, col = 'red')
lines(m1_conc ~ time, data = out11_3, col = 'orange')
lines(m1_conc ~ time, data = out11_0, col = 'black')

plot(CH4 ~ time, data = out11_0, type = 'l')
plot(VSd ~ time, data = out11_0, type = 'l')
plot(m0_conc ~ time, data = out11_0, type = 'l')
lines(m0_conc ~ time, data = out11_1, type = 'l', col = 'red')
plot(m1_conc ~ time, data = out11_0, type = 'l')
plot(m2_conc ~ time, data = out11_0, type = 'l')
plot(VFA_conc ~ time, data = out11_0, type = 'l')
```


# Next. Hydrogenotrophic methanogens with CO2 and H2
Possible but not super logical -- we won't have a lot of H2 production without consumption

# Next. Other reaction products for substrates

Hydrolysis and fermentation can produce more than just VFAs if desired.
Here we'll include CO2 (of dubious value in my opinion).
The get_fstoich() helper function can help to create the matrix.

```{r}
#sub_forms <- c(cellulose = 'C6H10O5', protein = 'C4 H6.1 O1.2 N', lipids = 'C57 H104 O6')
#source('../R/supp.R')
#get_fstoich(sub_forms)
```

It automatically converts to the correct units.

```{r}
fsmat <- matrix(c(0, -0.0157, -0.1079,
                 1,  1,       1),
               nrow = 2,
               byrow = TRUE,
               dimnames = list(
                 c('CO2', 'VFA'),
                 c('cellulose', 'protein', 'lipids')))

fsmat
```

This fsmat (fermentation stoichiometry) matrix goes in `sub_pars` (although it technically could go in any `_pars` argument and work).


```{r}
sub_pars3 <- list(
  subs = c('cellulose', 'protein', 'lipids'),
  T_opt_hyd = c(default = 60),
  T_min_hyd = c(default = 0),
  T_max_hyd = c(default = 90),
  hydrol_opt = c(cellulose = 0.01, protein = 0.1, lipids = 1),
  sub_fresh = c(cellulose = 20, protein = 20, lipids = 20),
  sub_init = c(cellulose = 20, protein = 20, lipids = 20),
  fstoich = fsmat 
)
```

But if we want to include CO2, it should be in the microbial stoichoimetry matrix, and a component too.

```{r}
mstoich3
```

```{r}
grp_pars4 <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat2,
  mstoich = mstoich3
)
```


```{r}
inf_pars3
chem_pars2
```

```{r}
devtools::load_all()
out12 <- abmneo(
  50,
  storage = storage2,
  inf_pars = inf_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars3,
  chem_pars = chem_pars2
)
```

```{r}
plot(CH4 ~ time, data = out12, type = 'l')
lines(CO2 ~ time, data = out12, col = 'blue')
```
