---
title: 'Demo of `abmneo()` function in abmneo'
output: pdf_document
author: Sasha D. Hafner
date: "`r format(Sys.time(), '%d %B, %Y %H:%M')`"
---

# Overview
This demo shows:

# Prep

```{r}
devtools::load_all()
```

# Default input parameters

We need to set some parameters, first management parameters.
Here are some default sets, not always with plausible values.

Now pared-down mng_pars, only those that apply to both types ().

```{r}
mng_pars <- list(
  storage_depth = 2,     
  area = 100,              
  temp_C = 20,
  resid_enrich = 0.2 
)
```

Next substrate parameters, a new argument.
This defines substrates.
We could have any number with any names.
Note that hydrolysis uses CTM again (like anything here, that could be changed).

```{r}
sub_pars <- list(
  subs = c('VSd'),
  T_opt_hyd = c(VSd = 60),
  T_min_hyd = c(VSd = 0),
  T_max_hyd = c(VSd = 90),
  hydrol_opt = c(VSd = 0.1),
  sub_fresh = c(VSd = 50),
  sub_init = c(VSd = 50)
)
```

Microbial parameters are similar to other ABM versions, but inhibition is set separately now (and not included in this simple example).
But now we have microbial stoichiometry instead of the "special cases" that greatly complicate the code.
The ks par is now in a matrix, which could have more than one row (e.g., with sulfate reduction).

```{r}
mstoich <- matrix(
  c(-1, -1, -1, 1, 1, 1),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'CH4'),
    c('m0', 'm1', 'm2')
  )
)

mstoich

ksmat <- matrix(
  c(0.5, 0.5, 0.5),
  nrow = 1,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH'),
    c('m0', 'm1', 'm2')
  )
)

ksmat

```

```{r}
grp_pars <- list(
  grps = c('m0', 'm1', 'm2'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41),
  T_max = c(m0 = 25, m1 = 25, m2 = 38),
  ksmat = ksmat,
  mstoich = mstoich
)
```

The `dd_rate` parameter is for "death and decay".

These last two arguments are similar to other versions.
VFA is hard-coded as an intermediate and has its own elements.
A name is needed, and is `CH3COOH` here (and hard-coded for now).

```{r}
man_pars <- list(VFA_fresh = c(CH3COOH = 2), VFA_init = c(CH3COOH = 2), pH = 7, dens = 1000)
```

The chem_pars list now specifies which components are emitted as gases and which are otherwise solutes.
(All substrates are enriched--perhaps this should be an input too.)
Gases have no association with the slurry like solutes or substrates--their production is taken as emission and cumulative values are tracked.

```{r}
chem_pars <- list(COD_conv = c(CH4 = 5.32), gases = c('CH4'))
```

Conversion factor from:

```{r}
#> biogas::calcCOD('CH4') * molMass('CH4') / molMass('C')
#[1] 5.328893
```


# 1. Slurry mass series versus regular inputs
When it comes to how slurry level and the fill/empty schedule is entered, we now distinguish between *regular* inputs (as in repeating/consistent, as in earlier versions of `abmneo()`) and *series* inputs, where the schedule is inferred from time series data on slurry mass (what was called time-variable in earlier versions of `abmneo()`).
These are specified in the `storage` argument (to be renamed. . .  maybe simply `mass`???).
Mass dat for storage only has time and slurry_mass.
It is no longer a part of `var_pars`--the two are now independent (externally).

```{r}
mass_series <- data.frame(
  time = c(0, 1, 10, 15, 30), 
  slurry_mass = c(1000, 7000, 2000, 1500, 3000)
)

structure1 <- list(
  type = 'series',
  dat = mass_series
)

plot(slurry_mass ~ time, data = structure1$dat, type = 'o')
```

```{r}
devtools::load_all()
out1 <- abmneo(
  storage = structure1,
  50,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars
)
names(out1)
```

Unlike `ABM()`, no `var_pars` is needed.
And the fixed inputs like `slurry_prod_rate` that were ignored in these cases with `ABM()` are no longer expected.

```{r}
plot(slurry_mass ~ time, data = out1, type = 'l')
plot(CH4_emis_rate ~ time, data = out1, type = 'l')
plot(CH4 ~ time, data = out1, type = 'l')
plot(m0 ~ time, data = out1, type = 'l')
plot(CH3COOH ~ time, data = out1, type = 'l')
plot(VSd_conc ~ time, data = out1, type = 'l')
plot(m0_conc ~ time, data = out1, type = 'l')
plot(COD_load ~ time, data = out1, type = 'l')
```

With regular inputs, `slurry_prod_rate` etc. are now explicitly given.

```{r}
structure2 <- list(
  type = 'regular',
  slurry_prod_rate = 1E4, 
  slurry_mass = 1E4,     
  resid_depth = 0.1,      
  empty_int = 10,          
  wash_water = 100,            
  wash_int = NA,
  rest_d = 1
)
```

```{r}
devtools::load_all()
out2 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars
)
```
Q

```{r}
out2$slurry_mass
out2$slurry_mass_eff
```

```{r}
plot(slurry_mass ~ time, data = out2, type = 'l')
plot(CH4_emis_rate ~ time, data = out2, type = 'l')
plot(CH4 ~ time, data = out2, type = 'l')
plot(VSd ~ time, data = out2, type = 'l')
plot(CH3COOH ~ time, data = out2, type = 'l')
plot(m1 ~ time, data = out2, type = 'l')
```

# 2. Time-variable inputs
Time-variable inputs (now separate from slurry mass series data) can be combined with both of the above examples.
First create the var_pars object.

```{r}
temp_series <- data.frame(
  time = c(0, 1, 10, 15, 30), 
  temp_C = c(10, 20, 25, 15, 10)
)
var_pars <- list(var = temp_series)
```

Use with series data.

```{r}
devtools::load_all()
out3 <- abmneo(
  storage = structure1,
  50,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  var_pars = var_pars
)
```

```{r}
plot(temp_C ~ time, data = out3, type = 'l')
plot(CH4_emis_rate ~ time, data = out3, type = 'l')
```

Or regular inputs.

```{r}
devtools::load_all()
out4 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  var_pars = var_pars
)
```

```{r}
plot(slurry_mass ~ time, data = out4, type = 'l')
plot(temp_C ~ time, data = out4, type = 'l')
plot(CH4_emis_rate ~ time, data = out4, type = 'l')
```

# 3. Flexible substrates

Above only a single particulate substrate was used.
But any, with arbitrary names, can be defined.

```{r}
sub_pars2 <- list(
  subs = c('slow', 'mid', 'fast'),
  T_opt_hyd = c(default = 60),
  T_min_hyd = c(default = 0),
  T_max_hyd = c(default = 90),
  hydrol_opt = c(slow = 0.01, mid = 0.1, fast = 1),
  sub_fresh = c(slow = 20, mid = 20, fast = 20),
  sub_init = c(slow = 20, mid = 20, fast = 20)
)
```
Q

```{r}
devtools::load_all()
out5 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars,
  sub_pars = sub_pars2,
  chem_pars = chem_pars
)
```

```{r}
plot(fast_conc ~ time, data = out5, type = 'l', ylim = c(0, 100))
lines(slow_conc ~ time, data = out5, col = 'red')
lines(CH3COOH_conc ~ time, data = out5, col = 'blue')

plot(fast ~ time, data = out5, type = 'l', ylim = c(0, 5e6))
lines(slow ~ time, data = out5, col = 'red')
lines(CH3COOH ~ time, data = out5, col = 'blue')
```

This is the simplest way to specify substrates.
When entered in this way, there is no stoichiometry *per se*, but each g substrate COD produces 1 g CH3COOH intermediate.

# 4. Sulfate reduction and other metabolic pathways
The microbial metabolism code now used is completely flexibile with respect to reactants and products.
Sulfate reduction is not hard-coded but can easily and very explicitly be specified with the new approach.
We need the appropriate stoichiometry.
Masses are specified as:

1. COD mass, or if COD = 0,
2. N mass, or if N = 0,
3. C mass, or if C = 0,
4. S mass, or if S = 0,
5. total mass

So for SO4-2, we want the the mass of S divided by the COD of acetate, i.e., 32 / 64 = 0.5.
(We might think about flexibility in how stoichiometry is defined, e.g., allow user to input moles, but it is probably better to fix it.)

```{r}
mstoich2 <- matrix(
  c(-1, -1, -1, -1,
     0,  0,  0, -0.5,
     1,  1,  1,  0),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'SO4', 'CH4'),
    c('m0', 'm1', 'm2', 'sr1')
  )
)

mstoich2

ksmat2 <- matrix(
  c(0.5, 0.5, 0.5, 0.5,
    NA,   NA,  NA, 0.5),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'SO4'),
    c('m0', 'm1', 'm2',  'sr1')
  )
)

ksmat2
```

Put these in a grp_pars list.

```{r}
grp_pars2 <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat2,
  mstoich = mstoich2
)
```

Sulfate needs to be added as a component.

```{r}
man_pars2 <- list(VFA_fresh = c(CH3COOH = 2), 
		  VFA_init = c(CH3COOH = 2),
		  comps = c('SO4'),
                  comp_fresh = c(SO4 = 10), 
                  comp_init = c(SO4 = 10), 
                  pH = 7, 
		  dens = 1000)
```

```{r}
devtools::load_all()
out6 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars2,
  grp_pars = grp_pars2,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
)
```

```{r}
plot(slurry_mass ~ time, data = out6, type = 'l')
plot(temp_C ~ time, data = out6, type = 'l')
plot(CH4_emis_rate ~ time, data = out4, type = 'l')
lines(CH4_emis_rate ~ time, data = out6, type = 'l', col = 'red')
plot(CH4 ~ time, data = out6, type = 'l')
plot(SO4_conc ~ time, data = out6, type = 'l')
plot(sr1 ~ time, data = out6, type = 'l')
plot(m1 ~ time, data = out6, type = 'l')
plot(CH3COOH_conc ~ time, data = out4, type = 'l')
lines(CH3COOH_conc ~ time, data = out6, type = 'l', col = 'red')
```

Notice that the sulfate reduction reaction is not balanced.
Neither is methanogenesis, for that matter.
Additional products can be added pretty easily (and would be included in default parameter sets presumably).

Let's add H2S and CO2.
Note that the coefficient on CO2 is g C, and g S on H2S and SO4.

```{r}
mstoich3 <- matrix(
  c(-1,  -1,  -1,  -1,
     0,   0,   0,  -0.5,
     1.5, 1.5, 1.5, 3,
     0,   0,   0,   0.5,
     1,   1,   1,   0),
  nrow = 5,
  byrow = TRUE,
  dimnames = list(
    c('CH3COOH', 'SO4', 'CO2', 'H2S', 'CH4'),
    c('m0', 'm1', 'm2', 'sr1')
  )
)

mstoich3
```

```{r}
grp_pars3 <- list(
  grps = c('m0', 'm1', 'm2', 'sr1'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.05),
  xa_init = c(default = 0.05),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2, sr1 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28, sr1 = 25),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41, sr1 = 0),
  T_max = c(m0 = 25, m1 = 25, m2 = 38, sr1 = 40),
  ksmat = ksmat2,
  mstoich = mstoich3
)
```

Now we need more components.

```{r}
man_pars3 <- list(VFA_fresh = c(CH3COOH = 2), 
		  VFA_init = c(CH3COOH = 2),
		  comps = c('SO4', 'CO2', 'H2S'),
                  comp_fresh = c(SO4 = 10, CO2 = 1, H2S = 0), 
                  comp_init = c(SO4 = 10, CO2 = 1, H2S = 0), 
                  pH = 7, 
		  dens = 1000)
```

And we should specify that H2S and CO2 are gases, otherwise they accumulate like solutes.

```{r}
chem_pars2 <- list(COD_conv = c(CH4 = 3.01), gases = c('CH4', 'CO2', 'H2S'))
```

Remember that all this could be added to default parameter sets!

```{r}
devtools::load_all()
out7 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
)
```

```{r}
plot(CO2 ~ time, data = out7, type = 'l', col = 'blue')
lines(CH4 ~ time, data = out7, type = 'l')
lines(CH4 ~ time, data = out6, type = 'l', col = 'orange', lty = 2)
lines(H2S ~ time, data = out7, type = 'l', col = 'yellow')
plot(SO4_conc ~ time, data = out7, type = 'l')
plot(sr1 ~ time, data = out7, type = 'l')
plot(m1 ~ time, data = out7, type = 'l')
plot(CH3COOH_conc ~ time, data = out4, type = 'l')
lines(CH3COOH_conc ~ time, data = out7, type = 'l', col = 'red')
```

Repeat without any sulfate for sulfate reducers.

```{r}
man_pars4 <- list(VFA_fresh = c(CH3COOH = 2), 
		  VFA_init = c(CH3COOH = 2),
		  comps = c('SO4', 'CO2', 'H2S'),
                  comp_fresh = c(SO4 = 0, CO2 = 1, H2S = 0), 
                  comp_init = c(SO4 = 0, CO2 = 1, H2S = 0), 
                  pH = 7, 
		  dens = 1000)
```


```{r}
devtools::load_all()
out8 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars4,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
)
```
Q

```{r}
plot(CH4 ~ time, data = out8, type = 'l')
lines(CO2 ~ time, data = out8, type = 'l', col = 'blue')
lines(CH4 ~ time, data = out6, type = 'l', col = 'orange', lty = 2)
lines(H2S ~ time, data = out8, type = 'l', col = 'yellow')
```


Stoichiometry is all sorted out based on the chemical formula in `forms`, including how much VFA (if any) is produced and whether there is N mineralization (but then TAN needs to be added as a solute).




#mass_structure = list(
#  type = 'series',
#  data = 
#
#slurry_prod_rate = 10000, 
#                 slurry_mass = 1000,     
#                 resid_depth = 0.1,      
#                 empty_int = 100,          
#                 wash_water = 0,            
#                 wash_int = NA,
#                 rest_d = 0,


# Note
The "late" and "mid" options are still available, but now through `ctrl_pars`.
Here we can change the value through `add_pars`

# 5. Time-variable inputs part 2
Almost any model parameter can now be varied over time through var_pars.
If it is a single value without named elements for any time, it can be combined in the var data frame.
This excludes anything having to do with emptying, loading rate, or slurry level--if any of those need to vary non-regularly over time, the storage argument must be used!
So, that leaves probably only pH and temperature.
For example, we could simulate acidification with this.

```{r}
var_dat <- data.frame(
  time = c(0, 5, 10, 15, 30), 
  temp_C = c(10, 20, 25, 15, 10),
  pH = c(7, 6.5, 5, 5, 5)
)
var_pars2 <- list(var = var_dat)
```

```{r}
devtools::load_all()
out9 <- abmneo(
  50,
  storage = structure2,
  mng_pars = mng_pars,
  man_pars = man_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
  var_pars = var_pars2
)
```

```{r}
plot(slurry_mass ~ time, data = out9, type = 'l')
plot(temp_C ~ time, data = out9, type = 'l')
plot(pH ~ time, data = out9, type = 'l')
plot(CH4 ~ time, data = out9, type = 'l')
lines(H2S ~ time, data = out9, type = 'l', col = 'brown')
```

For other inputs that are set in named elements, make a separate data frame for each other argument with column names that give the element names.
For example, for changing substrate concentration,

```{r}
sub_fresh_dat = data.frame(time = c(0, 5, 10, 15, 30), 
                       VSd = c(50, 100, 0, 0, 200))
```

and a sudden increase in sulfate, 

```{r}
comp_fresh_dat = data.frame(time = c(0, 15, 30), 
                       SO4 = c(0, 5, 5))
```

or even a change in microbial parameters,

```{r}
qhat_opt_dat = data.frame(
  time = c(0, 22), 
  m0   = c(1, 0.1),
  m1   = c(1, 0.1),
  m2   = c(1, 0.1),
  sr1  = c(2, 1)
)
```

Then combine them all in a list, using the parameter element names for element names (e.g., `sub_fresh` is the name of an element in `sub_pars`).

```{r}
var_pars3 <- list(var = var_dat, sub_fresh = sub_fresh_dat, comp_fresh = comp_fresh_dat, qhat_opt = qhat_opt_dat)
```

Note that times do not need to be the same!
The data frames are merged and missing values are interpolated or filled forward (we need to align the language for this and mid etc.)

```{r}
devtools::load_all()
out10 <- abmneo(
  storage = structure2,
  50,
  mng_pars = mng_pars,
  man_pars = man_pars3,
  grp_pars = grp_pars3,
  sub_pars = sub_pars,
  chem_pars = chem_pars2,
  var_pars = var_pars3
)
```

```{r}
plot(slurry_mass ~ time, data = out10, type = 'l')
plot(CH4_emis_rate ~ time, data = out10, type = 'l')
plot(VSd_conc ~ time, data = out10, type = 'l')
plot(SO4_conc ~ time, data = out10, type = 'l')
plot(CH3COOH_conc ~ time, data = out5, type = 'l')
```

That is a lot of flexibility, I think.

# X. The startup argument

Works as before.
This example is based on out2

```{r}
structure3 <- list(
  type = 'regular',
  slurry_prod_rate = 1E4, 
  slurry_mass = 5E4,     
  resid_depth = 0.5,      
  empty_int = 30,          
  wash_water = 100,            
  wash_int = NA,
  rest_d = 1
)
```

Reduce initial methanogen concentrations and set fresh to zero.

```{r}
grp_pars4 <- list(
  grps = c('m0', 'm1', 'm2'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.01),
  xa_init = c(default = 0.01),
  dd_rate = c(default = 0.02),
  ksv = c(default = 1),
  qhat_opt = c(m0 = 1, m1 = 1, m2 = 2),
  T_opt = c(m0 = 18, m1 = 18, m2 = 28),
  T_min = c(m0 = 0, m1 = 6.41, m2 = 6.41),
  T_max = c(m0 = 25, m1 = 25, m2 = 38),
  ksmat = ksmat,
  mstoich = mstoich
)
```

```{r}
devtools::load_all()
out11_1 <- abmneo(
  90,
  delta_t = 0.1,
  storage = structure3,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 1
)
```


```{r}
out11_0 <- abmneo(
  90,
  delta_t = 0.1,
  storage = structure3,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 0
)

devtools::load_all()
out11_1 <- abmneo(
  90,
  delta_t = 0.1,
  storage = structure3,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 1
)

out11_2 <- abmneo(
  90,
  delta_t = 0.1,
  storage = structure3,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 2
)

out11_3 <- abmneo(
  90,
  delta_t = 0.1,
  storage = structure3,
  mng_pars = mng_pars,
  man_pars = man_pars,
  grp_pars = grp_pars4,
  sub_pars = sub_pars,
  chem_pars = chem_pars,
  startup = 10 
)

```

```{r}
out11_0$slurry_mass
out11_0$CH4_emis_rate
plot(slurry_mass ~ time, data = out11_0, type = 'l')

plot(CH4_emis_rate ~ time, data = out11_2, type = 'l', col = 'blue', ylim = c(0, 1.4E6))
lines(CH4_emis_rate ~ time, data = out11_1, col = 'red')
lines(CH4_emis_rate ~ time, data = out11_3, col = 'orange')
lines(CH4_emis_rate ~ time, data = out11_0, col = 'black')

plot(m1_conc ~ time, data = out11_2, type = 'l', col = 'blue', ylim = c(0, 0.06))
lines(m1_conc ~ time, data = out11_1, col = 'red')
lines(m1_conc ~ time, data = out11_3, col = 'orange')
lines(m1_conc ~ time, data = out11_0, col = 'black')

plot(CH4 ~ time, data = out11_0, type = 'l')
plot(VSd ~ time, data = out11_0, type = 'l')
plot(m0_conc ~ time, data = out11_0, type = 'l')
lines(m0_conc ~ time, data = out11_1, type = 'l', col = 'red')
plot(m1_conc ~ time, data = out11_0, type = 'l')
plot(m2_conc ~ time, data = out11_0, type = 'l')
plot(CH3COOH_conc ~ time, data = out11_0, type = 'l')
```


# Next. Hydrogenotrophic methanogens with CO2 and H2
Possible but not super logical -- we won't have a lot of H2 production without consumption
