---
title: 'Getting started with the `abmneo()` function'
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Getting started with the abmneo() function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: Sasha D. Hafner
date: "`r format(Sys.time(), '%d %B, %Y %H:%M')`"
---

# abmneo
The abmneo package, with a single `abmneo()` function, is an implementation of the ABM model for anaerobic biodegradation of organic wastes (Dalby et al., 2021).
The ABM model is also available in an earlier [ABM package](https://github.com/AU-BCE-EE/ABM).
The new package was developed as a simpler but more flexible alternative.
For user-defined substrates, microbial groups, or metabolic pathways, the newer package provides much more flexibility.
But it does not include all the functionality available in the ABM package, including rest days and washing between animal batches and evaporation of water. 
This vignette demonstrates the use of the `abmneo()` model function. 

# What's different?
The main differences between abmneo and ABM are:

1. package complexity and hard-coding,
2. handling of time-variable inputs, and
3. input flexibility for substrates, microbial groups, and stoichiometry.

In abmneo, time-variable inputs are treated using steps instead of the interpolation approach used in ABM.
This approach ostensibly reduces the accuracy of the model as a tradeoff for simpler code, more flexibility in which inputs can vary, and faster evaluation.
But in practice the effects seem negligible beyond a visual effect in plots.

Hard-coded behavior and special cases in ABM includes the microbial community, with fixed symbols for methanogen and sulfate reducer names and designated code for each. 
In contrast, the microbial community in abmneo is defined completely by input parameters, and stoichiometry is explicitly specified in input arguments.
Internally, shorter vectorized or matrix operations do the work for all groups without special cases.
Substrates and inhibition are two other areas where these types of major changes were made.

C++ code and the Rcpp package were incorporated into the ABM package for speed, but also bring some complexity.
The abmneo package uses only R, with any speed penalty apparently offset by other simplifications.
In the end abmneo is not obviously slower than ABM; it is even slightly faster in some cases.
Other dependencies were also eliminated in abmneo.

Some of the abmneo functionality is demonstrated in the examples below.
Not all the processes available in abmneo are shown however; respiration, ammonia production, and volatilization are available but not shown.

Let's load the new package and get started.

```{r}
library(abmneo)
```

# Input arguments and parameters
Here are the `abmneo()` arguments:

```{r, eval=FALSE}
abmneo <- function(
  storage,
  days = 365,
  delta_t = 1,
  times = NULL,
  inf_pars = NULL,
  grp_pars = NULL,
  sub_pars = NULL,
  chem_pars = list(COD_conv = c(CH4 = 5.32), gases = c('CH4', 'CO2', 'H2S')),
  ctrl_pars = list(
    rem_align_method = 'early',
    var_fill_method = 'interp',
    par_key = '\\.',
    h_rate_max_temp = 313
  ),
  var_pars = list(var = NULL),
  add_pars = NULL,
  pars = NULL,
  startup = 0,
  starting = NULL,
  warn = TRUE
) 
```

We'll come back to the `storage` argument later.
The `days` (total simulation time), `delta_t` (time step *in output*), and `times` (specific times to be included in output) are similar in ABM and straightforward.
Let's work on the `*_pars` arguments.
These all contain input parameters.
Here we will construct a set roughly similar to the ABM set `*_pars_pig2.0` default sets in ABM.

`inf_pars` has influent or inflow properties.
VFA is a hard-coded intermediate, and we need to set fresh and initial concentrations (g/kg) along with pH and density.
The presence of any other solutes, with any names, is set in the `comps` element, with concentrations in g/kg given in `comp_fresh` and `comp_init` elements.
Here the only optional solute we'll add is sulfate, which is needed for sulfate reduction.

```{r}
inf_ref <- list(
  VFA_fresh = 2,
  VFA_init = 2,
  comps = c('SO4'),
  comp_fresh = c(SO4 = 0.2),
  comp_init = c(SO4 = 0.2),
  pH = 7,
  dens = 1000
)
```

`grp_pars` defines the microbial community. 
The argument should have these elements:

* `grps`: group names
* `yield`: biomass yields on COD basis
* `xa_fresh`: active biomass concentrations in fresh influent
* `xa_init`: initial concentrations
* `d_max`: maximum death/decay rate (1/d)
* `qhat_opt`: maximum substrate untilization rate (g COD / g COD - d)
* `T_opt`: temperature of maximum activity (deg. C)
* `T_min`: minimum temperature with any activity (deg. C)
* `T_max`: maximum temperature with any activity (deg. C)
* `ksmat`: matrix of Monod parameters
* `mstoich`: matrix of stoichiometry

Definition of the microbial community is very flexible; any number of groups with any names and practically any metabolism can be specified.
Here we'll include those groups used in the latest ABM parameter set.
Let's start with the stoichiometry matrix.

```{r}
mstoich <- matrix(
  c(
    -1, -1, -1, -1, -1, -1, -1,
     0,  0,  0,  0,  0,  0, -0.50156,
     1,  1,  1,  1,  1,  1,  0
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4', 'CH4'),
    c('m0', 'm1', 'm2', 'm3', 'm4', 'm5', 'sr1')
  )
)
```

It is easier to understand by looking at the matrix with row and column names:

```{r}
mstoich
```

Columns are groups, showing the group that carries out the reaction, rows are substrates or products.
The values of 1 are an expression of chemical oxygen demand (COD) conservation--they show that each methanogen group produces 1 g of methane per g of VFA consumed (both on a COD basis, when biomass production is neglected).
Compounds without a COD are expressed as mass N (if present), S (if present), or otherwise total compound mass.
Here then `'SO4-2'` is as mass of S.
So where did the `0.50` come from?
The [micstoich package](https://github.com/AU-BCE-EE/micstoich) was used to sort out the stoichiometry of sulfate reduction.

```{r, eval = FALSE}
library(micstoich)
rxn <- micstoich('CH3COOH', acceptor = 'SO4-2', product = 'H2S')
rxn
molmass('SO4-2', 'S') * rxn['SO4-2'] / (calcCOD('CH3COOH', per = 'mol') * rxn['CH3COOH'])
```

This takes acetic acid as the representative VFA.
And it gives:

```{r, eval = FALSE}
> library(micstoich)
> rxn <- micstoich('CH3COOH', acceptor = 'SO4-2', product = 'H2S')
> rxn
CH3COOH   SO4-2      H+     H2S     HS-     H2O     CO2 
-0.1250 -0.1250 -0.1875  0.0625  0.0625  0.2500  0.2500 
> molmass('SO4-2', 'S') * rxn['SO4-2'] / (calcCOD('CH3COOH', per = 'mol') * rxn['CH3COOH'])
    SO4-2 
0.5015625 
```

Note that these coefficients do not account for microbial biomass--this information is entered as a microbial yield, and the stoichiometry matrix is updated and expanded internally.

We need one more matrix for `grp_pars`: `ksmat`, which has the Monod half-velocity parameters.
Here the values are from `mic_pars2.0` in the ABM repo.

```{r}
ksmat <- matrix(
  c(1.15, 1.15, 1.15, 1.15, 1.15, 1.15, 0.46,
      NA,   NA,   NA,   NA,   NA,   NA, 0.00694),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c('VFA', 'SO4'),
    c('m0', 'm1', 'm2', 'm3', 'm4', 'm5', 'sr1')
  )
)

ksmat
```

Note that `NA` values for sulfate and the methanogens.

Now let's fill in a `grp_pars` list.
Unlike the ABM package, there is no `all` keyword in abmneo, but the `default` keyword behaves the same way where no group is expicitly specified.

```{r}
grp_ref <- list(
  grps = c('m0', 'm1', 'm2','m3','m4','m5','sr1'),
  yield = c(default = 0.05, sr1 = 0.065),
  xa_fresh = c(default = 0.06),
  xa_init = c(default = 0.1),
  d_max = c(default = 0.02),
  qhat_opt =  c(m0 = 0.463, m1 = 0.746, m2 = 0.901, m3 = 2.80, m4 = 4.66, m5 = 7.46, sr1 = 8.34),
  T_opt = c(m0 = 18, m1 = 18,  m2 = 28,  m3 = 36, m4 = 44, m5 = 55, sr1 = 44),
  T_min = c(m0 =  0, m1 =  9.7, m2 = 9.7, m3 = 15, m4 = 26, m5 = 30, sr1 =  0),
  T_max = c(m0 = 25, m1 = 25,  m2 = 38,  m3 = 45, m4 = 51, m5 = 60, sr1 = 51),
  ksmat = ksmat,
  mstoich = mstoich
)
```

And check it.

```{r}
grp_ref
```

Substrates are completely defined in the `sub_pars` input.
Like microbial groups, any number and any names could be used.
Here we'll continue with something like the ABM parameters.
Hydrolysis kinetics follows the same exponential response as in the Arrhenius equation used in ABM, but the parameters are different
In abmneo, they are `h_rate_ref`: the rate constant (1/d) at 20 degrees C, and `h_rate_q10`: the relative change in rate over 10 degrees C.

```{r}
sub_ref <- list(
  subs = c('starch', 'Cfat', 'CPs', 'CPf', 'RFd'),
  sub_enrich = c('starch', 'Cfat', 'CPs', 'RFd'),
  xd = 'xd',
  h_rate_ref = c(starch = 0.029, Cfat = 0.058, CPs = 0.012, CPf = 0.012, RFd = 0.0058, xd = 0.0058),
  h_rate_q10 = c(default = 2.1),
  sub_fresh = c(starch = 5.2, Cfat = 28, CPs = 12, CPf = 9.5, RFd = 25, xd = 0),
  sub_init  = c(starch = 5.2, Cfat = 28, CPs = 12, CPf = 9.5, RFd = 25, xd = 0)
)
```

So now we've created lists for `inf_pars`, `grp_pars`, and `sub_pars`.
These are the only required parameter groups.

The `chem_pars` argument has COD conversion ratios in g COD per g output or input units.
These are needed for the COD balance and expressing output in useful units.
The default element is for methane, as g COD per g methane C.
VFA and substrates are all in COD units anyway, so there is no conversion.
The `gases` element names the components that should be treated as gases; they do not accumulate in the slurry.
Cumulative emission and emission rate is calculated for each gas.
All other components are solutes, which accumulate until the storage is emptied.

The `var_pars` argument is used for time-variable inputs.
We won't use it in the first example.
And `add_pars` is for adding or modifying parameters set through other arguments.
We won't use that in this first example either.

But we do need to sort out the `storage` argument.
There are two basic approaches that can be used: "regular", with a fixed rate of influent addition and a fixed emptying interval; and "series", where influent addition and storage emptying is completely flexible, and defined by a slurry level over time.

## Regular option

For the regular option, `abmneo()` needs a list with these elements:

```{r}
stor_reg_ref <- list(
  type = 'regular',
  slurry_prod_rate = 5, 
  storage_depth = 2,     
  area = 0.65,              
  temp_C = 20,
  resid_enrich = 0.6,
  slurry_mass = 6.5,     
  resid_depth = 0.05,      
  empty_int = 28
)
```

These are approximate values for slurry in a pig house with monthly emptying, with values for one pig.
Units here are kg for mass, m for length, and d for time.
Unlike the ABM package, washing of a slurry storage, e.g., between animal batches, is not incorporated.
(But it could be more manually incorporated with the `starting` argument--see the final example.)

## Series option

And for a series, the main element is a data frame with the measurements.

```{r}
mass_series <- data.frame(
  time =         c(0, 121, 122, 152, 153, 182, 214, 244, 365),
  slurry_mass = c(67,  80,   2,  13,   0.01,  11,  22,  32,  67)
)
plot(slurry_mass ~ time, data = mass_series, type = 'o')

stor_ser_ref <- list(
  type = 'series',
  dat = mass_series,
  storage_depth = 2,     
  area = 0.7,              
  temp_C = 18,
  resid_enrich = 0.6 
)
```

This is meant to represent outdoor storage tank, with values for single pig.
More typically, measured values (and more of them) will be read in and used.

# Running the model

Finally, let's run the model, first with the regular option.

```{r}
out_reg <- abmneo(
  storage = stor_reg_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref
)
```

Output is a data frame that contains gas emission rates, cumulative emission, and masses and concentrations of microbial biomass, substrates, VFA, and any other solutes.
The column names, and number of columns, depends on the inputs; each microbial group and each substrate has two associated columns (total mass in g COD and concentration in g COD / kg).

```{r}
names(out_reg)
```

The plots below show some results.

Methane emission rate, in g/d.

```{r}
plot(CH4_emis_rate ~ time, data = out_reg, type = 'l')
```

And cumulative emission in g.

```{r}
plot(CH4 ~ time, data = out_reg, type = 'l')
```

Microbial biomass (g COD).

```{r}
matplot(
  out_reg$time, 
  out_reg[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1:length(nn), type = 'l',
  ylab = 'Xa'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

And microbial biomass concentrations (g COD / kg).

```{r}
matplot(
  out_reg$time, 
  out_reg[, nn <- paste0(grp_ref$grps, '_conc')], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1:length(nn), 
  type = 'l',
  ylab = 'Xa conc'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

Substrate concentrations.

```{r}
matplot(
  out_reg$time, 
  out_reg[, nn <- paste0(sub_ref$subs, '_conc')], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'red'), 
  lty = 1:length(nn), 
  type = 'l',
  ylab = 'Substrate conc.'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

And VFA mass (g COD) and concentration (g COD / kg).

```{r}
plot(VFA ~ time, data = out_reg, type = 'l')
plot(VFA_conc ~ time, data = out_reg, type = 'l')
```

Of course most of the temporal patterns are driven by slurry mass, which is also returned.

```{r}
plot(slurry_mass ~ time, data = out_reg, type = 'l')
```

And so is slurry depth (m).

```{r}
plot(slurry_depth ~ time, data = out_reg, type = 'l')
```

And the series input option.

```{r}
out_ser <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref
)
```

Here are the same plots that were shown above for regular inputs.

```{r}
plot(CH4_emis_rate ~ time, data = out_ser, type = 'l')
```

```{r}
plot(CH4 ~ time, data = out_ser, type = 'l')
```

```{r}
matplot(
  out_ser$time, 
  out_ser[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1:length(nn), type = 'l',
  ylab = 'Xa'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

```{r}
matplot(
  out_ser$time, 
  out_ser[, nn <- paste0(grp_ref$grps, '_conc')], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1:length(nn), 
  type = 'l',
  ylab = 'Xa conc'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

```{r}
matplot(
  out_ser$time, 
  out_ser[, nn <- paste0(sub_ref$subs, '_conc')], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'red'), 
  lty = 1:length(nn), 
  type = 'l',
  ylab = 'Substrate conc.'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

```{r}
plot(VFA ~ time, data = out_ser, type = 'l')
plot(VFA_conc ~ time, data = out_ser, type = 'l')
```

```{r}
plot(slurry_mass ~ time, data = out_ser, type = 'l')
```

# Time-variable inputs

Most inputs can be varied over time.
They go into the `var_pars` argument as named data frames in a list.
Here is an example with temperature.

```{r}
temp_dat <- data.frame(time = seq(0, 365, length.out = 53))
temp_dat$temp_C <- 10 + 10 * sin((temp_dat$time - 130) / 365 * 2 * pi)
plot(temp_C ~ time, data = temp_dat, type = 'o')
```

```{r}
out_var <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  var_pars = list(var = temp_dat)
)
```

And some of the same plots again.

Emission rate.

```{r}
plot(CH4_emis_rate ~ time, data = out_var, type = 'l')
```

Cumulative emission.

```{r}
plot(CH4 ~ time, data = out_var, type = 'l')
```

Temperature is also returned.

```{r}
plot(temp_C ~ time, data = out_var, type = 'l')
```

Here you can see how time-varying inputs are handled.
Instead of smooth interpolation, they are held constant within the steps implied in the `var_pars` input.

# Default digeste parameter set

A single default parameter set for stored digestate is included with the first release of abmneo.

They include give microbial groups:

```{r}
grp_pars_dig_01
```

There are no optional solutes included, which means no sulfate reduction.
Fresh VFA concentration is 200 mg COD / kg.

```{r}
inf_pars_dig_01
```

There is a single grouped substrate `PS`, for "particulate substrate".
It is similar to the VS-based approach in ABM, but the name `VS` is was deemed confusing for a COD-based quantity (it typically refers to the total mass of solids lost during calcination at 550 degrees C).

```{r}
sub_pars_dig_01
```

We need `storage` inputs to use these default values.
This is an example included with the package.

```{r}
storage_ex_dig_01
```

Let's use all these in a model run.

```{r}
out_dig <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_dig_01,
  sub_pars = sub_pars_dig_01
)
```

```{r}
plot(CH4 ~ time, data = out_dig, type = 'l')
plot(slurry_mass ~ time, data = out_dig, type = 'l', ylim = c(0, 2e6))
```

# Optional arguments

This section briefly presents the use of some of the optional `abmneo()` arguments.

The optional `add_pars` argument can be used to change individual parameters without repeating all the values in an argument.
For example, in `ctrl_pars` there is a `rem_align_method` parameter that determines the interpretation of slurry removal from a series `storage` object.
It could be changed from the default with calls like these.

```{r}
out_dig_early <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_dig_01,
  sub_pars = sub_pars_dig_01,
  add_pars = list(rem_align_method = 'early')
)
```

```{r}
out_dig_late <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_dig_01,
  sub_pars = sub_pars_dig_01,
  add_pars = list(rem_align_method = 'late')
)
```

```{r}
plot(slurry_mass ~ time, data = out_dig, type = 'l', ylim = c(0, 2e6))
lines(slurry_mass ~ time, data = out_dig_early, col = 'red')
lines(slurry_mass ~ time, data = out_dig_late, col = 'blue')
```

This `rem_align_method` argument will have a tiny effect when slurry mass inputs are close together in time.
But when available measurements are sparse, as in this example dataset, it can have a large effect.

```{r}
plot(CH4 ~ time, data = out_dig, type = 'l')
lines(CH4 ~ time, data = out_dig_early, col = 'red')
lines(CH4 ~ time, data = out_dig_late, col = 'blue')
```

It is important to make sure the selected approach best matches the measurements.
For example, if level measurements were made just after emptying, the `'late'` approach should be used.

For changing a parameter that is nested, like the initial concentration of `PS`, there is a `.` notation that has been build into the function.

```{r}
out_dig_psi <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_dig_01,
  sub_pars = sub_pars_dig_01,
  add_pars = list(sub_init.PS = 0)
)
```

```{r}
plot(PS_conc ~ time, data = out_dig, type = 'l', ylim = c(0, 15))
lines(PS_conc ~ time, data = out_dig_psi, col = 'red')
```

The `startup` argument can be used to run the specified simulation some number of times.
This simulates, e.g., years of operation and gives the microbial community time to reach, if not a true steady-state, then a state that is not highly sensitive to starting conditions.


```{r}
out_dig_1s <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_dig_01,
  sub_pars = sub_pars_dig_01,
  startup = 1
)
```

```{r}
out_dig_3s <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_dig_01,
  sub_pars = sub_pars_dig_01,
  startup = 3
)
```

```{r}
plot(CH4 ~ time, data = out_dig, type = 'l')
lines(CH4 ~ time, data = out_dig_1s, col = 'red')
lines(CH4 ~ time, data = out_dig_3s, col = 'blue')
```

Here is an example with time-variable temperature.

```{r}
out_var_1s <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  var_pars = list(var = temp_dat),
  startup = 1
)
```

```{r}
out_var_3s <- abmneo(
  storage = stor_ser_ref,
  365,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  var_pars = list(var = temp_dat),
  startup = 3
)
```

```{r}
plot(CH4_emis_rate ~ time, data = out_var, type = 'l')
lines(CH4_emis_rate ~ time, data = out_var_1s, col = 'red')
lines(CH4_emis_rate ~ time, data = out_var_3s, col = 'blue', lty = 2)
```

The `times` argument returns specific times only.
For example, for (roughly) monthly output:

```{r}
out_var_sel <- abmneo(
  storage = stor_ser_ref,
  365,
  times = seq(31, 365, length.out = 12),
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  var_pars = list(var = temp_dat)
)
```

The normal, instantaneous (point) and cumulative, results are the same regardless of which times are returned.

```{r}
plot(CH4_emis_rate ~ time, data = out_var, type = 'l')
lines(CH4_emis_rate ~ time, data = out_var_sel, type = 'o', pch = 19, col = 'red')
```

```{r}
plot(CH4 ~ time, data = out_var, type = 'l')
lines(CH4 ~ time, data = out_var_sel, type = 'o', pch = 19, col = 'red')
```

But there is another output column `CH4_emis_rate_ave` that is not.
This column has the average rate over the preceding interval.

```{r}
plot(CH4_emis_rate_ave ~ time, data = out_var, type = 'S')
lines(CH4_emis_rate_ave ~ time, data = out_var_sel, type = 'S', pch = 19, col = 'red')
```

The `times` argument also works for "regular" storage inputs.

```{r}
out_reg_sel <- abmneo(
  storage = stor_reg_ref,
  365,
  times = seq(31, 365, length.out = 12),
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref
)
```

```{r}
plot(CH4_emis_rate_ave ~ time, data = out_reg, type = 'S')
lines(CH4_emis_rate_ave ~ time, data = out_reg_sel, type = 'S', col = 'red')
```

In this case, it is much easier to see the slight increase in emission rate over time.
This simulation would benefit from a startup period!

The `starting` argument is used to feed the results from one simulation into another one.
The only results transfered over are concentrations--of substrates, all solutes, and microbial biomass.
With the ability to change any inputs, this argument provides a lot of flexiblity.

For example, we can simulate 3 emptying cycles, a rest period, washing (by 10x dilution of slurry), and then another 3 months.
First the 3 cycles:

```{r}
out_reg_1 <- abmneo(
  storage = stor_reg_ref,
  3 * 28,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref
)
```

Here is slurry mass over that time:

```{r}
out_reg_1$slurry_mass
```

Let's keep that same final slurry mass in the pit for a few months. 
We'll use `add_pars` to change the initial slurry mass and slurry addition rate here, but we could also create a modified `storage` object.
Let's change the temperature as well, just for fun.
Notice the simplest way to indicate that emptying does not occur, with the `empty_int` argument (a value of `0` would have the same effect).

```{r}
out_reg_2 <- abmneo(
  storage = stor_reg_ref,
  3 * 30,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  add_pars = list(slurry_mass = 146.5, slurry_prod_rate = 0, empty_int = 0, temp_C = 16),
  starting = out_reg_1
)
```

We can combine the output from the two cosecutive simulations, but any cumulative value will be incorrect in the second one.

```{r}
out_reg_2$time <- out_reg_2$time + max(out_reg_1$time)
out_comb <- rbind(out_reg_1, out_reg_2)
```

```{r}
plot(CH4_emis_rate ~ time, data = out_comb, type = 'l')
```

What drives that increase in emission rate?
There is no new substrate added.
Slurry mass should be constant after the end of the first simulation.

```{r}
plot(slurry_mass ~ time, data = out_comb, type = 'l')
```

But take a look at what happens to the microbial community when slurry removal stops.

```{r}
matplot(
  out_comb$time, 
  out_comb[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1:length(nn), type = 'l',
  ylab = 'Xa'
)
legend('topright', legend = nn, col = cc, lty = 1:length(nn))
```

We could edit the output from one simulation before using it in the `starting` argument of the next.
This could be used to simulate, e.g., washing.

The `var_pars` example above doesn't show everything that argument can do. 
As an example, let's kill off methanogen groups `m0` and `m1` in the `out_reg_2` simulation after 50 d and see what happens to methane production.
Here are the original results.

```{r}
matplot(
  out_reg_2$time, 
  out_reg_2[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1, type = 'l',
  ylab = 'Xa'
)
legend('topright', legend = nn, col = cc, lty = 1)
```

```{r}
out_reg_2k <- abmneo(
  storage = stor_reg_ref,
  3 * 30,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  add_pars = list(slurry_mass = 146.5, slurry_prod_rate = 0, empty_int = 0, temp_C = 16),
  var_pars = list(qhat_opt = data.frame(time = c(0, 50), m0 = c(0.463, 0), m1 = c(0.746, 0))),
  starting = out_reg_1
)
```

Let's add to the plot from above.

```{r}
matplot(
  out_reg_2$time, 
  out_reg_2[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1, type = 'l',
  ylab = 'Xa'
)
legend('topright', legend = nn, col = cc, lty = 1)

matlines(
  out_reg_2k$time, 
  out_reg_2k[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'),
  lty = 3
)
legend('topright', legend = nn, col = cc, lty = 3)
```

To combine this type of `var_pars` input with a simple scalar parameter, just put the scalar in a `var` data frame, like this:

```{r}
out_reg_2kt <- abmneo(
  storage = stor_reg_ref,
  3 * 30,
  inf_pars = inf_ref,
  grp_pars = grp_ref,
  sub_pars = sub_ref,
  add_pars = list(slurry_mass = 146.5, slurry_prod_rate = 0, empty_int = 0, temp_C = 16),
  var_pars = list(var = data.frame(time = c(0, 60), temp_C = c(16, 5)), qhat_opt = data.frame(time = c(0, 50), m0 = c(0.463, 0), m1 = c(0.746, 0))),
  starting = out_reg_1
)
```

```{r}
matplot(
  out_reg_2$time, 
  out_reg_2[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'), 
  lty = 1, type = 'l',
  ylab = 'Xa'
)
legend('topright', legend = nn, col = cc, lty = 1)

matlines(
  out_reg_2kt$time, 
  out_reg_2kt[, nn <- grp_ref$grps], 
  col = cc <- c('blue', 'orange', 'green', 'black', 'black', 'black', 'red'),
  lty = 4
)
legend('topright', legend = nn, col = cc, lty = 3)
```

The last optional feature covered here is inhibition.
Two matrices are required--one with the highest concentration causing no effect and one with the lowest concentration causing complete inhibition.
To use inhibition we need to create a complete `grp_pars` object.
Here we'll do that with only 3 microbial groups, for less typing.

```{r}
grp_pars_inhib <- list(
  grps = c('m1', 'm2','m3'),
  yield = c(default = 0.05),
  xa_fresh = c(default = 0.06),
  xa_init = c(default = 0.06),
  d_max = c(default = 0.02),
  qhat_opt =  c(m1 = 0.5, m2 = 1, m3 = 2),
  T_opt = c(m1 = 18, m2 = 28, m3 = 36),
  T_min = c(m1 =  2, m2 =  2, m3 = 15),
  T_max = c(m1 = 25, m2 = 38, m3 = 45),
  ksmat = matrix(
    c(1.1, 1.1, 1.1),
    nrow = 1,
    byrow = TRUE,
    dimnames = list(
      c('VFA'),
      c('m1', 'm2', 'm3')
    )
  ),
  mstoich = matrix(
    c(
      -1, -1, -1,
       1,  1,  1
    ),
    nrow = 2,
    byrow = TRUE,
    dimnames = list(
      c('VFA', 'CH4'),
      c('m1', 'm2', 'm3')
    )
  ),
  ic0 = matrix(
    c(8, 8, 8,
      8, 8, 8,
      6, 6, 6),
    nrow = 3,
    byrow = TRUE,
    dimnames = list(
      c('VFA_conc', 'pH', 'pH'),
      c('m1', 'm2', 'm3')
    )
  ), ic100 = matrix(
    c(15, 15, 15,
      9, 9, 9,
      5, 5, 5),
    nrow = 3,
    byrow = TRUE,
    dimnames = list(
      c('VFA_conc', 'pH', 'pH'),
      c('m1', 'm2', 'm3')
    )
  )
)
```

Notice how pH is entered twice and reversed?
That makes it inhibit activity both below 6 and above 8.
Inhibition is complete below 5 and above 9.
We'll use the example and default inputs for other arguments.

```{r}
out_inhib_1 <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_inhib,
  sub_pars = sub_pars_dig_01
)
```

```{r}
plot(CH4_emis_rate ~ time, data = out_inhib_1, type = 'l')
plot(pH ~ time, data = out_inhib_1, type = 'l')
plot(VFA_conc ~ time, data = out_inhib_1, type = 'l')
```

The inhibitors never get close to the ic0 even.
So we'll increase them with `var_pars`.

```{r}
out_inhib_2 <- abmneo(
  storage = storage_ex_dig_01,
  365,
  inf_pars = inf_pars_dig_01,
  grp_pars = grp_pars_inhib,
  sub_pars = sub_pars_dig_01,
  var_pars = list(var = data.frame(time = c(0, 100), pH = c(7.5, 5.2)))
)
```

```{r}
plot(CH4_emis_rate ~ time, data = out_inhib_1, type = 'l')
lines(CH4_emis_rate ~ time, data = out_inhib_2, type = 'l', col = 'red')
```

There is our inhibition.

# References

Dalby, F.R., Hafner, S.D., Petersen, S.O., Vanderzaag, A., Habtewold, J., Dunfield, K., Chantigny, M.H., Sommer, S.G., 2021. A mechanistic model of methane emission from animal slurry with a focus on microbial groups. PLOS ONE 16, e0252881. 10.1371/journal.pone.0252881

